---
title: "Import, Trim, and Denoise ASVs"
author: "Sarah Tanja"
date: 09/03/2026
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = FALSE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "", # Prevents appending '##' to beginning of lines in code output
  dev = "jpeg",
  dpi = 600,
  fig.retina = 2)        
```

# Background

I'm re-running the commands already run by Adam Waalkes of the Salipante Lab (qiime2 commands found in `../salipante/Sarah_StonyCoral/250405_QIIME2_filtered_StonyCloral_final.txt`), the original commands I re-use and/or modify are also block-quoted prior to executable code chunks in this document.

# Load packages

```{r}
library(knitr)
library(dplyr)
library(stringr)
library(readr)
library(reticulate)
```

# Activate qiime2 environment in R

This is equivalent to `conda activate qiime2-2023.5` in the terminal but allows running qiime commands in r chunks with engine ='bash' throughout the quarto markdown doc.

```{r load-qiime-env, eval=TRUE}
use_condaenv(condaenv = "qiime2-2023.5", conda = "/home/shared/8TB_HDD_02/stanja/miniconda3/condabin/conda")

# Check successful env loading
py_config()
```

# Sequence files

-   Here I am working with raw `.fastq.gz` files located in the `../data/` directory
-   There are two `fastq.gz` files for each of the 63 samples, 1 forward read and 1 reverse read (R1 & R2)

# Metadata

-   metadata for the samples are located at `../metadata/metadata.tsv`

```{r}
metadata <- read_tsv("../metadata/metadata.tsv")
```

```{r, engine='bash'}
qiime metadata tabulate \
  --m-input-file ../metadata/metadata.tsv \
  --o-visualization ../output/denoise/metadata.qzv
```

# Modify read manifest

-   the read_manifest file the salipante lab used is here:

```{r}
read_manifest <- read_tsv("../salipante/241121_StonyCoral/250414_StonyCoral_read_manifest.tsv")

head(read_manifest)
```

The read_manifest is a list of the sample_id's and their corresponding forward and reverse fastq.gz file paths. In this directory, the fastq.gz files are located at `../data/`

However the paths must be absolute in order to run `qiime tools import` in the next step. Let's modify the read_manifest so that it works in this directory structure.

```{r}
normalizePath("../data/")
```

```{r, eval=TRUE}
read_manifest_mod <- read_manifest %>% 
  mutate(`forward-absolute-filepath` = str_replace(`forward-absolute-filepath`,
                                                  "^/mnt/labs/salipante/data/sequencing_runs/241121_StonyCoral/",
                                                  "/media/4TB_JPG_ext/stanja/gitprojects/coral-embryo-microbiome/data/"),
         `reverse-absolute-filepath` = str_replace(`reverse-absolute-filepath`,
                                                   "^/mnt/labs/salipante/data/sequencing_runs/241121_StonyCoral/",
                                                   "/media/4TB_JPG_ext/stanja/gitprojects/coral-embryo-microbiome/data/")
  )

head(read_manifest_mod)
```

```{r}
write_tsv(read_manifest_mod, "../input/read_manifest.tsv")
```

# Import data to QIIME

> `qiime tools import --type 'SampleData[PairedEndSequencesWithQuality]' --input-path 250414_StonyCoral_read_manifest.tsv --output-path 250414_StonyCoral_imported_reads.qza --input-format PairedEndFastqManifestPhred33V2 Imported 250414_StonyCoral_read_manifest.tsv as PairedEndFastqManifestPhred33V2 to 250414_StonyCoral_imported_reads.qza`

```{r, engine='bash'}
qiime tools import \
  --type 'SampleData[PairedEndSequencesWithQuality]' \
  --input-path ../input/read_manifest.tsv \
  --output-path ../output/denoise/imported_reads.qza \
  --input-format PairedEndFastqManifestPhred33V2 
```

# Trim

> `#klind primers so: qiime cutadapt trim-paired --i-demultiplexed-sequences 250414_StonyCoral_imported_reads.qza --p-cores 20 --p-front-f CCTACGGGNGGCWGCAG --p-front-r GACTACHVGGGTATCTAATCC --verbose --p-discard-untrimmed   --o-trimmed-sequences 250414_StonyCoral_imported_reads_trimmed.qza`

```{r, engine='bash'}
qiime cutadapt trim-paired \
  --i-demultiplexed-sequences ../output/denoise/imported_reads.qza \
  --p-cores 20 \
  --p-front-f CCTACGGGNGGCWGCAG \
  --p-front-r GACTACHVGGGTATCTAATCC \
  --verbose \
  --p-discard-untrimmed \
  --o-trimmed-sequences ../output/denoise/imported_reads_trimmed.qza
```

```{r, engine='bash'}
qiime tools peek ../output/denoise/imported_reads_trimmed.qza
```

## Create qzv of trimmed reads

> `qiime demux summarize --i-data 250414_StonyCoral_imported_reads_trimmed.qza --o-visualization 250414_StonyCoral_imported_reads_trimmed.qzv`

```{r, engine='bash'}
qiime demux summarize \
  --i-data ../output/denoise/imported_reads_trimmed.qza \
  --o-visualization ../output/denoise/imported_reads_trimmed.qzv
```

Use QIIME 2 View to view the `.qzv` files:

1.  Copy the `.qzv` or file to your local machine via <https://github.com/sarahtanja/coral-embryo-microbiome/blob/main/output/>

    Open <https://view.qiime2.org> in a browser.

    Drag and drop your file there to view it.

# Check quality scores

> `#qualities seemed fine, running across a number of trims to see which is best (qiime2-amplicon-2023.9) tron:/mnt/labs/salipante/data/projects/241121_StonyCoral/270x200 $ qiime dada2 denoise-paired --i-demultiplexed-seqs ../250414_StonyCoral_imported_reads_trimmed.qza --p-n-threads 20 --p-trunc-len-f 270 --p-trunc-len-r 200 --verbose --o-table 250414_StonyCoral_270x200_featureTable.qza --o-representative-sequences 250414_StonyCoral_240x230_representative-sequences.qza --o-denoising-stats 250414_StonyCoral_240x230_denoising-stats.qza`

```{r, engine='bash'}
qiime dada2 denoise-paired \
  --i-demultiplexed-seqs ../output/denoise/imported_reads_trimmed.qza \
  --p-n-threads 20 \
  --p-trunc-len-f 270 \
  --p-trunc-len-r 200 \
  --verbose \
  --o-table ../output/denoise/feature_table.qza \
  --o-representative-sequences ../output/denoise/representative_sequences.qza \
  --o-denoising-stats ../output/denoise/denoising_stats.qza
```

# Filter representative sequences

# Reverse compliment

## Export representative sequences from QIIME2

```{r}
qiime tools export \
  --input-path rep-seqs.qza \
  --output-path rep_seqs_exported
```

> \# 4 other read lengths
>
> qiime feature-table summarize \--i-table 250414_StonyCoral_270x200_featureTable.qza \--o-visualization 250414_StonyCoral_270x200_featureTable.qzv \--m-sample-metadata-file ../250416_StonyCoral_metadata.tsv
>
> qiime feature-table tabulate-seqs \--i-data 250414_StonyCoral_270x200_representative-sequences.qza \--o-visualization 250414_StonyCoral_270x200_representative-sequences.qzv
>
> qiime metadata tabulate \--m-input-file 250414_StonyCoral_270x200_denoising-stats.qza \--o-visualization 250414_StonyCoral_270x200_denoising-stats.qzv
>
> qiime diversity alpha-rarefaction \--i-table 250414_StonyCoral_270x200_featureTable.qza \--p-min-depth 10 \--p-steps 30 \--p-max-depth 400000 \--m-metadata-file ../250416_StonyCoral_metadata.tsv \--o-visualization 250416_StonyCoral_270x200_alpha_rarefaction_curves.qzv

```{r, engine='bash'}
qiime feature-table summarize \
  --i-table ../output/feature_table.qza \
  --o-visualization ../output/feature_table.qzv \
  --m-sample-metadata-file ../metadata/metadata.tsv
```

```{r, engine='bash'}
qiime feature-table tabulate-seqs \
  --i-data ../output/representative_sequences.qza \
  --o-visualization ../output/representative_sequences.qzv
```

```{r, engine='bash'}
qiime metadata tabulate \
  --m-input-file ../output/denoising_stats.qza \
  --o-visualization ../output/denoising_stats.qzv
```

> \# looked at the curves and 270x200 was best as usual, and 400K looked good

# Summary & next steps
