---
title: "Running PICRUSt2"
author: "Sarah Tanja"
date: 2026-03-16
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = FALSE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "") # Prevents appending '##' to beginning of lines in code output)        
```

# Background

[How does PICRUSt2 work? What does it do?](https://github.com/picrust/picrust2/wiki/How-does-PICRUSt2-work%3F-What-does-it-do%3F)

> PICRUSt2 aims to predict the functional content of microbial communities based on amplicon/marker gene sequences. Very simply, PICRUSt2 takes information about which microbes are present in a sample and predicts what these microbes might be doing. PICRUSt2 compares the marker gene sequences from a sample to known genomes from similar microbes in order to predict the genes that are likely present and what functions they might perform.

-   fasta formatted representative sequences (ASVs)

-   feature table (tab-delimited) containing abundance of ASVs within samples

::: callout-note
Some advice from Dr. Steve Salipante

> Yes it is reasonable to look at picrust output and see if there are functional shifts, but it is not clear how adequate picrust libraries are to allow metabolic inference of these particular populations.  I would recommend that the actual analysis of metabolic shifts be conducted in Maaslin, as before, to account for covariates.
>
> it is possible to give sarah the input files she would need for the analysis. \[see [https://forum.qiime2.org/t/picrust2-input-files-for-bacteria/27968](https://urldefense.com/v3/__https://forum.qiime2.org/t/picrust2-input-files-for-bacteria/27968__;!!K-Hz7m0Vt54!iZIXAiB2i2QtliEgrFMztEKZHQZuAILNAY54WW_klnDOn0rCNia6tnE-GrFq0efQ8IeJkkAwEJWtIM0$)\]. For the record, you could isolate the ASV table from DADA2 in qiime (these are stored in rep-seqs-dada2.qza, according to [https://forum.qiime2.org/t/how-to-generate-an-asv-table-from-dada2-output/9973](https://urldefense.com/v3/__https://forum.qiime2.org/t/how-to-generate-an-asv-table-from-dada2-output/9973__;!!K-Hz7m0Vt54!iZIXAiB2i2QtliEgrFMztEKZHQZuAILNAY54WW_klnDOn0rCNia6tnE-GrFq0efQ8IeJkkAwpWyxhvY$)), and convert it to biom format \[[https://forum.qiime2.org/t/how-to-generate-an-asv-table-from-dada2-output/9973/2](https://urldefense.com/v3/__https://forum.qiime2.org/t/how-to-generate-an-asv-table-from-dada2-output/9973/2__;!!K-Hz7m0Vt54!iZIXAiB2i2QtliEgrFMztEKZHQZuAILNAY54WW_klnDOn0rCNia6tnE-GrFq0efQ8IeJkkAwc-L0gtk$)\]. 
>
> having said this it may just be easier to use the picrust2 plugin for qiime and hand over those outputs, since we already have the inputs ready in qiime
>
> [https://github.com/picrust/picrust2/wiki/q2-picrust2-Tutorial](https://urldefense.com/v3/__https://github.com/picrust/picrust2/wiki/q2-picrust2-Tutorial__;!!K-Hz7m0Vt54!iZIXAiB2i2QtliEgrFMztEKZHQZuAILNAY54WW_klnDOn0rCNia6tnE-GrFq0efQ8IeJkkAwVcNVfvQ$)
>
> [https://github.com/picrust/picrust2/wiki/Manually-install-QIIME-2-plugin](https://urldefense.com/v3/__https://github.com/picrust/picrust2/wiki/Manually-install-QIIME-2-plugin__;!!K-Hz7m0Vt54!iZIXAiB2i2QtliEgrFMztEKZHQZuAILNAY54WW_klnDOn0rCNia6tnE-GrFq0efQ8IeJkkAwAif4ef0$)\
> \
> id say generate those outputs and then run them through maaslin to look for significant feature abundances. -- S.S.
:::

# Setup

# Load libraries

```{r, eval=TRUE}
library(tidyverse)
library(qiime2R)
```

# Activate picrust2 conda env in Terminal
```
conda activate picrust2
```

# PICRUSt2 qiime pipeline

```{r, engine='bash'}
qiime picrust2 full-pipeline --help
```

> The required inputs are --i-table and --i-seq, which need to correspond to QIIME 2 artifacts of types FeatureTable\[Frequency\] and FeatureData\[Sequence\], respectively. The Feature Table needs to contain the abundances of ASVs (i.e. a BIOM table) and the sequence file needs to be a FASTA file containing the sequences for each ASV.

> These files correspond to the ASV count table, ASV sequences, and metadata for the samples.

## Feature table (BIOM)

located at path `../../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_featureTable_filtered.qza`

```{r, engine='bash', eval=TRUE}
qiime tools peek ../../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_featureTable_filtered.qza
```

::: callout-important
DO NOT SUBSET YOUR ASVs! 
Once we pass the full set of ASVs and their relative abundance counts to PICRUSt2, the software will use the full set of ASVs to predict the functional content of the microbial communities for every sample! If you subset your ASVs to only those that map to the taxa that were found to be differentially abundant/prevalent in the maaslin3 analysis, the follow-up maaslin3 analysis (after running PICRUSt2) will be biased because the functional predictions will have been made based on a subset of the ASVs that were present in the samples. This would lead to an unfair comparison between the taxa-level maaslin3 results and the function-level maaslin3 results, because the function-level results would be based on a biased subset of the data.
:::

```{r}
feat_tbl <- read_qza("../../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_featureTable_filtered.qza")$data
```

## Sequence file

located at path `../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_representative-sequences.qza`

```{r, engine='bash', eval=TRUE}
qiime tools peek ../../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_representative-sequences.qza

```

Filter representative sequences

```{r}
rep_seqs <- read_qza("../../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_representative-sequences.qza")$data
```

## Run `qiime picrust2 full-pipeline`

```{r, engine='bash'}
qiime picrust2 full-pipeline \
   --i-table ../../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_featureTable_filtered.qza \
   --i-seq ../../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_representative-sequences.qza \
   --output-dir ../../output/picrust2v2 \
   --p-threads 4 \
   --p-max-nsti 2 \
   --verbose
```

::: callout-warning
Standard error of the above failed command: Warning - 1169 input sequences aligned poorly to reference sequences (--min_align option specified a minimum proportion of 0.8 aligning to reference sequences).
:::

See [issue#122](https://github.com/picrust/picrust2/issues/122)

> There's a good chance your input sequences are on the negative strand and aren't being aligned properly. In case some one may need this, use seqtk to get reverse complementary seqs, then using picrust2_pipeline.py. seqtk seq -r otus.fa.format.fasta \> otus_rc.fa

### Install SEQTK

Run in terminal and proceed with 'y'

```         
conda install -c bioconda seqtk
```

### Reverse compliment

Representative sequences (.qza → .fasta)

```{r, engine='bash'}
qiime tools export --help
```

```{r, engine='bash'}
qiime tools export \
  --input-path ../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_representative-sequences.qza \
  --output-path ../output/

```

Reverse compliment the sequences with `seqtk`

```{r, engine='bash'}
seqtk seq -r ../output/dna-sequences.fasta > ../output/rep_seqs_rc.fasta
```

```{r, engine='bash'}
qiime tools list-types
```

```{r, engine='bash'}
qiime tools import \
  --type 'FeatureData[Sequence]' \
  --input-path ../output/rep_seqs_rc.fasta \
  --output-path ../output/rep_seqs_rc.qza \
  --input-format DNAFASTAFormat
```

```{r, engine='bash'}
qiime tools peek ../output/rep_seqs_rc.qza
```

### Run picrust with reverse compliment

```{r, engine='bash'}
qiime picrust2 full-pipeline \
   --i-table ../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_featureTable_filtered.qza \
   --i-seq ../output/rep_seqs_rc.qza \
   --output-dir ../output/picrust2 \
   --p-placement-tool sepp \
   --p-threads 4 \
   --p-hsp-method pic \
   --p-max-nsti 2 \
   --verbose
```

::: callout-warning
```         
Standard error of the above failed command:
Stopping - all 7182 input sequences aligned poorly to reference sequences (--min_align option specified a minimum proportion of 0.8 aligning to reference sequences).
```

OK this likely indicates that input sequences were originally on the positive strand, and the seqtk reverse compliment we made is the negative strand.. the number of input sequences that failed to align increased from 1169/7182 (16%) to 7182/7182 (100%)
:::

> can also consider lowering the --min_align flag so reads have a better chance of matching poorly characterized taxa.

--p-min-align PROPORTION Range(0.0, 1.0) Proportion of the total length of an input query sequence that must align with reference sequences. Any sequences with lengths below this value after making an alignment with reference sequences will be excluded from the placement and all subsequent steps. \[default: 0.8\]

## Lower --p-min-align

```{r, engine='bash'}
qiime picrust2 full-pipeline \
   --i-table ../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_featureTable_filtered.qza \
   --i-seq ../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_representative-sequences.qza \
   --output-dir ../output/picrust2 \
   --p-placement-tool sepp \
   --p-threads 4 \
   --p-hsp-method pic \
   --p-max-nsti 2 \
   --verbose \
   --p-min-align 0.1
```

### Troubleshoot error

::: callout-warning
Error running this command: mv /tmp/tmpulv7acg9/picrust2_out/intermediate/place_seqs/sepp_out/output_placement.newick /tmp/tmpulv7acg9/picrust2_out/out.tre

Standard error of the above failed command: mv: relocation error: /lib/x86_64-linux-gnu/libacl.so.1: symbol getxattr version ATTR_1.0 not defined in file libattr.so.1 with link time reference
:::

The error indicates there are multiple or corrupted versions between `libacl.so.1` and `libattr.so.1` on your system. To get around this we can set the library path when running qiime commands.

## Set `LD_LIBRARY_PATH` when running QIIME

```{r, engine='bash'}
LD_LIBRARY_PATH=$CONDA_PREFIX/lib:$LD_LIBRARY_PATH \
qiime picrust2 full-pipeline \
   --i-table ../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_featureTable_filtered.qza \
   --i-seq ../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_representative-sequences.qza \
   --output-dir ../output/picrust2 \
   --p-placement-tool sepp \
   --p-threads 4 \
   --p-hsp-method pic \
   --p-max-nsti 2 \
   --verbose \
   --p-min-align 0.1
```

> ::: callout-warning
> Warning - 661 input sequences aligned poorly to reference sequences (--min_align option specified a minimum proportion of 0.1 aligning to reference sequences). These input sequences will not be placed and will be excluded from downstream steps. 323 of 6521 ASVs were above the max NSTI cut-off of 2.0 and were removed from the downstream analyses.
> :::

The output artifacts of this command are the red boxes in the flowchart [here](https://huttenhower.sph.harvard.edu/picrust/#:~:text=PICRUSt%202.0%20Flowchart).

These output files in (q2-picrust2_output) are:

-   ec_metagenome.qza : EC metagenome predictions (rows are EC numbers and columns are samples).

-   ko_metagenome.qza : KO metagenome predictions (rows are KOs and columns are samples).

-   pathway_abundance.qza : MetaCyc pathway abundance predictions (rows are pathways and columns are samples).

The artifacts are all of type FeatureTable\[Frequency\], which means they can be used with QIIME 2 plugins that process and analyze these datatypes.

# Run with `mp` & `epa-ng`

> if users have at least 16 GB of RAM they should place reads with `epa-ng`

```         
--p-placement-tool TEXT Choices('epa-ng', 'sepp')
                       Placement tool to use when placing sequences into
                       reference tree. EPA-ng is the default, but SEPP is less
                       memory intensive.                   [default: 'epa-ng']
```

> **We recommend that in practice users use the `mp` method**

```         
--p-hsp-method TEXT Choices('mp', 'emp_prob', 'pic', 'scp',
    'subtree_average') Which hidden-state prediction method to use.
                                                               [default: 'mp']
```

```{r, engine='bash'}
LD_LIBRARY_PATH=$CONDA_PREFIX/lib:$LD_LIBRARY_PATH \
qiime picrust2 full-pipeline \
   --i-table ../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_featureTable_filtered.qza \
   --i-seq ../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_representative-sequences.qza \
   --output-dir ../output/picrust2_epa-ng \
   --p-placement-tool epa-ng \
   --p-threads 20 \
   --p-hsp-method mp \
   --p-max-nsti 2 \
   --verbose \
   --p-min-align 0.1
```

::: callout-warning
Warning - 661 input sequences aligned poorly to reference sequences (--min_align option specified a minimum proportion of 0.1 aligning to reference sequences). These input sequences will not be placed and will be excluded from downstream steps. 287 of 6521 ASVs were above the max NSTI cut-off of 2.0 and were removed from the downstream analyses.
:::

# Next up ..

## [**Analysis of PICRUSt2 output**](https://github.com/picrust/picrust2/wiki/How-does-PICRUSt2-work%3F-What-does-it-do%3F#:~:text=pipeline%20here.-,Analysis%20of%20PICRUSt2%20output,-You%20can%20essentially)

You can essentially treat these as you would any other microbiome feature table (like your results based on taxonomy within the samples), but this remains a bit of an open question. Some things that users are often interested in are:

-   Which taxa are predicted to contribute to a function within different samples

-   Whether predicted functional abundance profiles differ between samples from different treatments

-   Whether the predicted abundance of certain functions differs between samples from different treatments

Some potentially useful tools for this further analysis are (note that we are not involved in the development/maintenance of these tools unless noted):

-   [ggpicrust2](https://github.com/cafferychen777/ggpicrust2) - a comprehensive package for analysing PICRUSt2 output

-   [JarrVis](https://github.com/dhwanidesai/JarrVis) - for making sankey diagrams to visualise the taxa that contribute to different functions within different sample groupings (you can see an example of it being used for analysis of PICRUSt2 output [here](https://github.com/dhwanidesai/JarrVis/wiki/Reanalysis-of-functions-predicted-from-16S-amplicon-sequence-data-in-a-time-series-%E2%80%93-sea-water-samples-(Raes-et-al.,-2022)) (developed/maintained by a past member of the Langille lab)

-   [FuncDiv](https://academic.oup.com/bioinformatics/article/39/1/btac809/6909011?login=true) - R package for calculating the diversity of taxa that contribute to different functions (developed/maintained by Gavin Douglas, main developer of PICRUSt2 - although no longer involved in development/maintenance - and a past member of the Langille lab)

-   [Microbiome Helper 2](https://github.com/LangilleLab/microbiome_helper/wiki/Home-%E2%80%90-Microbiome-Helper-2) - this is currently a work in progress, but we plan to have some complete tutorials for the bioinformatic and statistical analysis of microbiome data, including PICRUSt2 output (developed/maintained by us in the Langille lab)

-   [WHAM!](https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-018-4870-z) - designed for visualisation of taxa/functions within metagenomic data but could also be used with taxa within samples along with PICRUSt2 output

-   [BURRITO](https://borenstein-lab.github.io/burrito/) - visualisation tool for exploring links between taxonomy and function within microbiome data

-   [MetaViz](https://epiviz.github.io/metaviz/) - another tool designed for visualisation and analysis of metagenomic sequencing data, but may work for PICRUSt2 output

```{r sessionInfo}
sessionInfo()
```
