---
title: "Compositional Data Analysis" 
subtitle: "Following Gloor et al. 2017 recommendations for microbiome compositional data analysis"
author: "Sarah Tanja" 
date: 10/17/2025 
date-modified: today
format: 
  gfm: 
    toc: true 
    number-sections: true
  html: 
    theme: journal
    highlight-style: github 
    page-layout: article 
    code-background: true
    code-tools: 
      source: true 
      toggle: true
    toc: true 
    toc-depth: 2
    toc-location: left 
    number-sections: true 
    df-print: kable 
    smooth-scroll: true 
    link-external-icon: true 
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
bibliography: "../../microbiome_bibtex.bib"
editor: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "" # Prevents appending '##' to beginning of lines in code output  
)
```

# Background

This analysis follows the recommendations from Gloor et al. 2017 for compositional data analysis of microbiome data.
Microbiome sequence count data are compositional - they represent relative abundances constrained to sum to a total (sequencing depth).
This requires specialized statistical methods that account for the compositional nature of the data.

Key recommendations from Gloor et al. 2017: - Use centered log-ratio (CLR) transformation to convert compositional data to Euclidean space - Calculate Aitchison distance (Euclidean distance on CLR-transformed data) for dissimilarity - Use variance-based methods like PCA on CLR-transformed data - Perform statistical tests on transformed data

# Setup

```{r}
# Define file paths
feature_table_path <- "../../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_featureTable_filtered.qza"
metadata_path <- "../../metadata/meta.csv"
taxonomy_path <- "../../salipante/241121_StonyCoral/270x200/250414_270x200_representative-sequences_taxonomy.qza"
```

## Load packages

```{r, eval=FALSE}
# Check R version
# R.version.string

# Install and load required packages
#if (!requireNamespace("BiocManager", quietly = TRUE)) {
#  install.packages("BiocManager")
#}
# Install Bioconductor packages
#if (!requireNamespace("phyloseq", quietly = TRUE)) {
#  BiocManager::install("phyloseq")
#}
# Install qiime2R from GitHub
#if (!requireNamespace("qiime2R", quietly = TRUE)) {
#  devtools::install_github("jbisanz/qiime2R")
#}
```

```{r}
# Load libraries
library(tidyverse)
library(qiime2R)
library(phyloseq)
library(vegan)
library(compositions)
library(zCompositions)
library(ggplot2)
library(reshape2)
library(corrplot)
library(ggsidekick) # for theme_sleek from sean anderson
library(patchwork)
library(ape)
```


## Custom ggplot themes

```{r theme}
theme_sleek_axe <- function() {
  theme_sleek() +
    theme(
      panel.border = element_blank(),
      axis.line.x  = element_line(color = "grey70"),
      axis.line.y  = element_line(color = "grey70"),
      
      plot.title = element_text(
        size  = 12,
        color = "grey30",
        hjust = 0.5,
        margin = margin(b = 20)
      ),
      
      strip.text = element_text(size = 10, color = "grey30"),
      
      plot.subtitle = element_text(
        size   = 10,
        color  = "grey30",
        hjust  = 0.5,
        margin = margin(t = 0)
      ),
      
      axis.title = element_text(size = 10, color = "grey30"),
      axis.text = element_text(size = 10, color = "grey30"),
      legend.title = element_text(size = 10, color = "grey30"),
      legend.text = element_text(size = 10, color = "grey30")
    )
}


theme_pptx <- function() {
  theme_sleek() +
    theme(
      panel.border = element_blank(),
      axis.line.x  = element_line(color = "grey70"),
      axis.line.y  = element_line(color = "grey70"),
      
      plot.title = element_text(
        size  = 40,
        color = "grey30",
        hjust = 0.5,
        margin = margin(b = 20)
      ),
      
      strip.text = element_text(size = 24, color = "grey30"),
      
      plot.subtitle = element_text(
        size   = 24,
        color  = "grey30",
        hjust  = 0.5,
        margin = margin(b = 40)
      ),
      
      axis.title = element_text(size = 24, color = "grey30"),
      axis.text = element_text(size = 24, color = "grey30"),
      legend.title = element_text(size = 24, color = "grey30"),
      legend.text = element_text(size = 24, color = "grey30")
    )
}

```

## Geoms

```{r geoms}
geom_defaults <- function() {
  update_geom_defaults("point",      list(size = 4, stroke = 1))
  update_geom_defaults("line",       list(linewidth = 0.5))
  update_geom_defaults("boxplot",    list(linewidth = 0.5))
  update_geom_defaults("violin",     list(linewidth = 0.5))
  update_geom_defaults("vline",      list(linewidth = 0.5))
  update_geom_defaults("hline",      list(linewidth = 0.5))
  update_geom_defaults("abline",     list(linewidth = 0.5))
  update_geom_defaults("errorbar",   list(linewidth = 0.5))
  update_geom_defaults("pointrange", list(linewidth = 0.5))
  update_geom_defaults("crossbar",   list(linewidth = 0.5))
  update_geom_defaults("linerange",  list(linewidth = 0.5))
  update_geom_defaults("path", list(linewidth = 1))
  update_geom_defaults("segment", list(linewidth = 0.5))
  update_geom_defaults("text", list(size = 3))
}

geom_pptx <- function() {
  update_geom_defaults("point",   list(size = 10, stroke = 4))
  update_geom_defaults("line",    list(linewidth = 1.5))
  update_geom_defaults("boxplot", list(linewidth = 1.2))
  update_geom_defaults("violin",  list(linewidth = 1.2))
  update_geom_defaults("vline", list(linewidth = 1.5))
  update_geom_defaults("hline", list(linewidth = 1.5))
  update_geom_defaults("abline", list(linewidth = 1.5))
  update_geom_defaults("errorbar", list(linewidth = 1.5))
  update_geom_defaults("pointrange", list(linewidth = 1.5))
  update_geom_defaults("crossbar", list(linewidth = 1.5))
  update_geom_defaults("linerange", list(linewidth = 4))
  update_geom_defaults("path", list(linewidth = 4))
  update_geom_defaults("segment", list(linewidth = 1.5))
  update_geom_defaults("text", list(size = 7))
}

```


## Set colorschemes

```{r colors}
leachate.colors <- c(control = "#5ABDAC", 
                low = "#BBC7FF",
                mid = "#7D8BFF", 
                high = "#592F7D")

stage.5.colors <- c(
  egg           = "#FFE066",
  cleavage      = "#FFBA08",
  morula        = "#F48C06",
  prawnchip     = "#DC2F02",
  earlygastrula = "#6A040F"
  )

stage.3.colors <- c(
  cleavage      = "#FFBA08",
  prawnchip     = "#DC2F02",
  earlygastrula = "#6A040F"
  )

night.colors <- c(July_6th = "#21918C", 
                  July_7th = "#578B21", 
                  July_8th = "#1E2440")

status.colors <- c(typical   = "#75C165", 
                   uncertain = "#E3FAA5", 
                   malformed = "#8B0069")
```

# Load Data

## Load metadata

```{r}
# Load metadata
meta <- read_csv(metadata_path)

# set factors
meta <- meta %>% 
  mutate(
    collection_date = as.Date(collection_date, format = "%d-%b-%Y"),
    stage    = factor(stage,    levels = c("cleavage", "prawnchip", "earlygastrula"), ordered = TRUE),
    leachate = factor(leachate, levels = c("control", "low", "mid", "high"),        ordered = TRUE),
    spawn_night = factor(
      collection_date,
      levels  = as.Date(c("06-Jul-2024", "07-Jul-2024", "08-Jul-2024"), format = "%d-%b-%Y"),
      labels  = c("July 6th", "July 7th", "July 8th"),
      ordered = TRUE
    )
  )


# View metadata structure
str(meta)
```

## Import QIIME2 artifacts

### Feature table

Our qiime2 feature table is in a .qza artifact format, which we can read into R using the `qiime2R` package.
The feature table contains the "count" data for each sample (in columns) and Amplicon Sequence Variance (ASV) feature (in rows).
The ASV features in the feature table rows are renamed as MD5 hashes of the DNA sequence.
Instead of storing the full sequence as the feature ID, the pipeline stores a hash of the sequence string.The hash string uniquely identifies one exact DNA sequence in the feature table.

ASV (Amplicon Sequence Variant) :

-   An exact biological sequence inferred after error correction.
-   No clustering threshold.
-   Each unique sequence (after denoising) is treated as its own unit.
-   Higher resolution and reproducible across studies.
-   Think exact sequence identity.

```{r}
# Load feature table from QIIME2 artifact
feature_table <- read_qza(feature_table_path)$data
```

### Taxonomy

The taxonomy table links the ASV MD5 hashes to the lowest taxonomic level

```{r}
# Load taxonomy (optional, for annotation)
taxonomy <- read_qza(taxonomy_path)$data

glimpse(taxonomy)
```

::: callout-important
Why are there still **"Unassigned"** ASVs?
I thought I had filtered those out...
should I do that now?
Or would that skew the sample similarity?
I think it could skew sample similarity analysis so I'll leave it in...
but flag this to ask Dr. Steve Salipante or Dr. Amy Van Cise ....
:::

```{r}
# Display data dimensions
cat("Feature table dimensions:", dim(feature_table), "\n")
cat("Number of samples in metadata:", nrow(meta), "\n")
```

## Prepare data for analysis

```{r}
# Convert feature table to matrix (samples as rows, features as columns)
# QIIME2 format typically has features as rows, samples as columns
feature_matrix <- as.matrix(feature_table)
feature_matrix_t <- t(feature_matrix)  # Transpose so samples are rows

# Ensure sample IDs match between feature table and metadata
sample_ids <- rownames(feature_matrix_t)
meta_filt <- meta %>%
  filter(sample_id %in% sample_ids) %>%
  arrange(match(sample_id, sample_ids))

cat("Samples in feature table:", nrow(feature_matrix_t), "\n")
cat("Samples in metadata:", nrow(meta_filt), "\n")
```

# CLR Transformation

The centered log-ratio (CLR) transformation converts compositional data to Euclidean space.
It handles zeros by replacing them with small values (geometric Bayesian multiplicative replacement).

```{r}
# Replace zeros with small values using cmultRepl from zCompositions
# This is necessary because log transformation is undefined for zeros
feature_matrix_nozero <- cmultRepl(feature_matrix_t, method = "CZM", label = 0)

# Apply CLR transformation
feature_clr <- clr(feature_matrix_nozero)

# Convert to data frame for easier handling
feature_clr_df <- as.data.frame(feature_clr)

write_csv(feature_clr_df, "../../output/CLR_feature_table.csv")


cat("CLR-transformed data dimensions:", dim(feature_clr_df), "\n")
head(feature_clr_df[, 1:5])
```

# Aitchison Distance is a Beta Diversity metric for compositional data

The Aitchison distance is the Euclidean distance between samples after CLR transformation.
This is the appropriate distance metric for compositional data.

```{r}
# Calculate Aitchison distance (Euclidean distance on CLR-transformed data)
aitchison_dist <- dist(feature_clr, method = "euclidean")

# Display distance matrix summary
cat("Aitchison distance matrix summary:\n")
summary(as.vector(aitchison_dist))
```

# PCA Biplots

Variance-based compositional PCA on CLR transformed feature table.

## Perform PCA

```{r}
# Perform PCA on CLR transformed feature table 
pca_result <- prcomp(feature_clr, scale. = FALSE)

# Calculate variance explained
variance_explained <- summary(pca_result)$importance[2, ] * 100

# Create data frame for plotting
pca_df <- data.frame(
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  sample_id = rownames(feature_clr)
)

# Merge with metadata
pca_df <- pca_df %>%
  left_join(meta_filt, by = "sample_id")
```

## Plot stage x leachate x cross

```{r, fig.height=6}
pca_plot_cross <- ggplot(
  pca_df,
  aes(
    x = PC1, y = PC2,
    fill = stage,                   # inside color
    colour = factor(parents),       # outline color
    shape = leachate                # shape must be 21–25
  )
) +
  coord_equal()+
  geom_point(size = 3.5, stroke = 1.5, alpha = 0.8) +   # stroke controls outline width
  scale_shape_manual(values = c(21, 22, 23, 24, 25)[seq_along(levels(pca_df$leachate))]) +
  # light grey -> dark grey for discrete stages (shows in legend)
  scale_fill_grey(name = "Stage", start = 0.1, end = 1) +
  labs(
    title = "PCA of CLR transformed  by Stage, Leachate, and Parental cross",
    x = paste0("PC1 (", round(variance_explained[1], 2), "% variance)"),
    y = paste0("PC2 (", round(variance_explained[2], 2), "% variance)"),
    shape = "Leachate",
    color = "Parental cross"
  ) +
  theme_sleek_axe() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  )+
  # 3) make the fill legend draw a filled shape with an outline
  guides(
    shape  = guide_legend(order = 1),
    fill   = guide_legend(override.aes = list(shape = 21, stroke = 1.1, colour = "black"), 
                          order = 2),
    colour = guide_legend(order = 3)
    
  ) 

pca_plot_cross

# Save plot
ggsave("../../output/compositional_pca_cross.png", pca_plot_cross, 
       width = 10, height = 7, dpi = 300)
```

## Plot stage x leachate x spawn night

```{r, fig.height=6}
pca_plot_spawn_night <- ggplot(
  pca_df,
  aes(
    x = PC1, y = PC2,
    fill = stage,                   # inside color
    colour = factor(spawn_night),       # outline color
    shape = leachate                # shape must be 21–25
  )
) +
  coord_equal()+
  geom_point(size = 3.5, stroke = 1.5, alpha = 0.8) +   # stroke controls outline width
  scale_shape_manual(values = c(21, 22, 23, 24, 25)[seq_along(levels(pca_df$leachate))]) +
  # light grey -> dark grey for discrete stages (shows in legend)
  scale_fill_grey(name = "Stage", start = 0.1, end = 1) +
  labs(
    title = "PCA of all ASVs by Stage, Leachate, and Spawn night",
    x = paste0("PC1 (", round(variance_explained[1], 2), "% variance)"),
    y = paste0("PC2 (", round(variance_explained[2], 2), "% variance)"),
    shape = "Leachate",
    color = "Spawn night"
  ) +
  theme_sleek_axe() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  )+
  # 3) make the fill legend draw a filled shape with an outline
  guides(
    shape  = guide_legend(order = 1),
    fill   = guide_legend(override.aes = list(shape = 21, stroke = 1.1, colour = "black"), 
                          order = 2),
    colour = guide_legend(order = 3)
    
  ) 

pca_plot_spawn_night

# Save plot
ggsave("../../output/compositional_pca_spawn_night.png", pca_plot_spawn_night, 
       width = 10, height = 7, dpi = 300)
```

## Interactive plotly

::: callout-note
This will only display in html format
:::

```{r, echo = knitr::is_html_output(), eval = knitr::is_html_output()}
library(ggplot2)
library(plotly)

# add a tooltip aesthetic to your existing ggplot
pca_plot_hpf <- ggplot(
  pca_df,
  aes(
    x = PC1, y = PC2,
    fill = stage,                    # inside color
    colour = factor(parents),        # outline color
    shape = leachate,                # shape must be 21–25
    text = paste0(
      "Stage: ", stage,
      "<br>Leachate: ", leachate,
      "<br>Parent: ", parents
    )
  )
) +
  coord_equal()+
  geom_point(size = 3.5, stroke = 1.5, alpha = 0.8) +
  scale_shape_manual(values = c(21, 22, 23, 24, 25)[seq_along(levels(pca_df$leachate))]) +
  scale_fill_grey(name = "Stage", start = 0.1, end = 1) +
  labs(
    title = "PCA of all OTUs by Stage, Leachate, and Parental cross",
    x = paste0("PC1 (", round(variance_explained[1], 2), "% variance)"),
    y = paste0("PC2 (", round(variance_explained[2], 2), "% variance)"),
    shape = "Leachate",
    color = "Parental cross"
  ) +
  theme_sleek_axe() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5)
  ) +
  guides(
    shape  = guide_legend(order = 1),
    fill   = guide_legend(override.aes = list(shape = 21, stroke = 1.1, colour = "black"),
                          order = 2),
    colour = guide_legend(order = 3)
  )

# convert to plotly and use only the custom text in the tooltip
pca_plot_hpf_interactive <- ggplotly(pca_plot_hpf, tooltip = "text")
pca_plot_hpf_interactive

```


# PERMANOVA

Here we run a PERMANOVA (Permutational Multivariate Analysis of Variance) on the Aitchison distance matrix to test for differences in microbial composition across experimental factors.

-   **Strata** are groups within which to constrain permutations.
    Using `strata = spawn night` controls for spawn night by constraining shuffles to within the spawn night groupings; this removes spawn night as a source of spurious differences.

-   **Dispersion sensitivity**: PERMANOVA can flag significance if groups have different dispersions (spread), not just different centroids.
    Consider checking with `betadisper` and `permutest`.

-   Unbalanced designs / small strata: With blocking, you need enough replicates within each strata and across factor levels to allow permutations; otherwise p-values become conservative or unavailable.
    **This is why we don't fit a betadispersion model to parental cross, there aren't enough replicates.**

## Fit betadispr

`betadisper` tests whether the within-group multivariate dispersion (spread) differs among groups.
It’s the multivariate analogue of Levene’s test: it compares the distances of samples to their group centroid.

What happens under the hood:

-   Performs a PCoA on your distance matrix.
-   Computes each group’s centroid in that ordination space.
-   Calculates each sample’s distance to its group centroid.
-   Stores those distances for downstream tests/plots.

Define groups to test

```{r}
grp_leachate <- meta_filt$leachate
grp_stage    <- meta_filt$stage     
grp_int    <- interaction(meta_filt$leachate, meta_filt$stage)
grp_night <- meta_filt$spawn_night
```

### night

```{r perm-test}
set.seed(02192026)
# Fit beta dispersion model 
bd_night <- betadisper(aitchison_dist, group = grp_night, type = "centroid")

# Permutation test with no blocking
perm_night <- permutest(bd_night, permutations = 999, pairwise = FALSE)

# Result
perm_night
```

::: callout-note
multivariate dispersion differs across spawn nights (groups = nights).
In other words, samples from some nights are more spread out around their night-specific centroid than others.

A permutation test for homogeneity of multivariate dispersions revealed a significant difference in dispersion among groups ( F2,60=6.29F_{2,60} = 6.29F2,60​=6.29, p=0.003p = 0.003p=0.003, 999 permutations), indicating that within-group variability in multivariate distance to centroid differed across nights.

betadisper with group = spawn_night significant heterogeneous dispersion among nights.
If your primary tests are leachate/stage, treat night as a nuisance: block by night in PERMANOVA (strata = spawn_night) so permutations happen within each night.
This keeps night-to-night variance/dispersion from inflating your p-values for treatment effects.
:::

```{r pca-plot}
plot(bd_night, hull = TRUE, ellipse = TRUE, label = FALSE, cex = 0.8)
```

In a betadisper object, the sample coordinates are in `bd_night$vectors` and the group centroids are in `bd_night$centroids`.
You can extract these for custom plotting or further analysis.
`chull()` is a function convex hull that finds the outermost points that form the smallest convex polygon around a set of points, which is useful for visualizing group dispersion in ordination plots, wrapped in `slice()` which keeps only the rows corresponding to those hull points for each group.


```{r night-pcoa, fig.width=6, fig.height=4}
# sample scores (vectors)
bd_night_df <- as.data.frame(bd_night$vectors[, 1:2])
bd_night_df$group <- bd_night$group

# centroid scores
bd_night_centroids <- as.data.frame(bd_night$centroids[, 1:2])
bd_night_centroids$group <- factor(rownames(bd_night_centroids), levels = levels(bd_night$group))

# axis labels with %variance explained
eig <- bd_night$eig
pct <- eig / sum(eig) * 100
xlab <- sprintf("PCoA 1 (%.1f%%)", pct[1])
ylab <- sprintf("PCoA 2 (%.1f%%)", pct[2])

# hulls
hulls <- bd_night_df %>%
  group_by(group) %>%
  slice(chull(PCoA1, PCoA2))

# force consistent factor levels across spawn night
unique(bd_night_df$group)
names(night.colors.alt)

lvls <- levels(factor(bd_night_df$group))

bd_night_df <- bd_night_df %>% mutate(group = factor(group, levels = lvls))
hulls <- hulls %>% mutate(group = factor(group, levels = lvls))
bd_night_centroids <- bd_night_centroids %>% mutate(group = factor(group, levels = lvls))

# plot
bd_night_plot <- ggplot(bd_night_df, aes(PCoA1, PCoA2, color = group)) +
                      geom_point(alpha = 0.85) +
                      geom_polygon(
                        data = hulls,
                        aes(fill = group),
                        alpha = 0.2,
                        color = NA,
                        show.legend = FALSE
                      ) +
                      geom_point(
                        data = bd_night_centroids,
                        shape = 4
                      ) +
                      stat_ellipse(level = 0.68) +
                      coord_equal() +
                     scale_x_continuous(
                      breaks = c(-30, 0, 30)
                    ) +
                    scale_y_continuous(
                      breaks = c(-20, 0, 20)
                    ) +
                      scale_color_manual(values = night.colors) +
                      scale_fill_manual(values = night.colors) +
                      theme_sleek_axe() +
                      labs(
                        title = "Spawn Night",
                        x = xlab,
                        y = ylab,
                        color = "spawn night"
                      )

geom_defaults()
bd_night_plot
ggsave(
  "../../figs/betadisp_night.png",
  width=6,
  height=4, 
  dpi=600,
  plot=bd_night_plot
)
```

```{r night-pptx, fig.width=16, fig.height=9}
geom_pptx()
betadisp_night_pptx <- bd_night_plot + theme_pptx()
betadisp_night_pptx
ggsave(
  "../../figs/betadisp_night_pptx.png",
  width=16,
  height=9,
  dpi=600,
  plot=betadisp_night_pptx
)
```

### leachate

```{r}
set.seed(02192026)
bd_leachate <- betadisper(aitchison_dist, group = grp_leachate, type = "centroid")

# Permutation test with optional blocking
perm_leachate <- permutest(bd_leachate, permutations = 999, pairwise = FALSE,
                  strata = meta_filt$spawn_night)  # optional
perm_leachate
```

::: callout-note
A permutation test for homogeneity of multivariate dispersions revealed no significant difference in dispersion across leachate (*F*~3,59~ = 0.1393, *p* = 0.939, 999 permutations)
:::

What this tests: Null hypothesis (Ho): all groups have equal average distance-to-centroid (equal dispersion/variance).If p \< 0.05, at least one group is more/less dispersed than another.
Why use the permutation test?
The distances depend on the PCoA, so permutation is safer than the plain ANOVA F-test.

```{r}
plot(bd_leachate, hull = TRUE, ellipse = TRUE, label = FALSE, cex = 0.8)
```

```{r leachate-pcoa}
# sample scores (vectors)
bd_leachate_df <- as.data.frame(bd_leachate$vectors[, 1:2])
bd_leachate_df$group <- bd_leachate$group

# centroid scores
bd_leachate_centroids <- as.data.frame(bd_leachate$centroids[, 1:2])
bd_leachate_centroids$group <- factor(rownames(bd_leachate_centroids), levels = levels(bd_leachate$group))

# axis labels with %variance explained
eig <- bd_leachate$eig
pct <- eig / sum(eig) * 100
xlab <- sprintf("PCoA 1 (%.1f%%)", pct[1])
ylab <- sprintf("PCoA 2 (%.1f%%)", pct[2])

# hulls
hulls <- bd_leachate_df %>%
  group_by(group) %>%
  slice(chull(PCoA1, PCoA2))

# force consistent factor levels across leachate
unique(bd_leachate_df$group)
names(leachate.colors)

lvls <- levels(factor(bd_leachate_df$group))

bd_leachate_df <- bd_leachate_df %>% mutate(group = factor(group, levels = lvls))
hulls <- hulls %>% mutate(group = factor(group, levels = lvls))
bd_leachate_centroids <- bd_leachate_centroids %>% mutate(group = factor(group, levels = lvls))


# plot
bd_leachate_plot <- ggplot(bd_leachate_df, aes(PCoA1, PCoA2, color = group)) +
                            geom_polygon(
                              data = hulls,
                              aes(fill = group),
                              alpha = 0.2,
                              color = NA,
                              show.legend = FALSE
                            ) +
                            geom_point(
                              data = bd_leachate_centroids,
                              shape = 4
                            ) +
                            stat_ellipse(level = 0.68) +
                            geom_point(alpha=0.85)+
                            coord_equal() +
                             scale_x_continuous(
                              breaks = c(-30, 0, 30)
                            ) +
                            scale_y_continuous(
                              breaks = c(-20, 0, 20)
                            ) +
                            scale_color_manual(values = leachate.colors) +
                            scale_fill_manual(values = leachate.colors) +
                            theme_sleek_axe() +
                            labs(
                              title = "Leachate",
                              x = xlab,
                              y = ylab,
                              color = "leachate"
                            )

geom_defaults()
bd_leachate_plot
ggsave(
  "../../figs/betadisp_leachate.png",
  width=6,
  height=4,
  dpi=600,
  plot=bd_leachate_plot
)
```

```{r leachate-pptx, fig.width=16, fig.height=9}
geom_pptx()
betadisp_leachate_pptx <- bd_leachate_plot + theme_pptx()
betadisp_leachate_pptx
ggsave(
  "../../figs/betadisp_leachate_pptx.png",
  width=16, 
  height=9,
  dpi=600,
  plot=betadisp_leachate_pptx
)
```

### stage

```{r}
set.seed(02192026)
bd_stage <- betadisper(aitchison_dist, group = grp_stage, type = "centroid")

# Permutation test with optional blocking
perm_stage <- permutest(bd_stage, permutations = 999, pairwise = FALSE)
                  #strata = meta_filt$spawn_night)  # optional
perm_stage
```

::: callout-note
A permutation test for homogeneity of multivariate dispersions revealed no significant difference in dispersion across stage (*F*~2,60~ = 1.0498, *p* = 0.359, 999 permutations)
:::

```{r}
plot(bd_stage, hull = TRUE, ellipse = TRUE, label = FALSE, cex = 0.8)
```

```{r stage-pcoa}

# sample scores (vectors)
bd_stage_df <- as.data.frame(bd_stage$vectors[, 1:2])
bd_stage_df$group <- bd_stage$group

# centroid scores
bd_stage_centroids <- as.data.frame(bd_stage$centroids[, 1:2])
bd_stage_centroids$group <- factor(rownames(bd_stage_centroids), levels = levels(bd_stage$group))

# axis labels with %variance explained
eig <- bd_stage$eig
pct <- eig / sum(eig) * 100
xlab <- sprintf("PCoA 1 (%.1f%%)", pct[1])
ylab <- sprintf("PCoA 2 (%.1f%%)", pct[2])

# hulls
hulls <- bd_stage_df %>%
  group_by(group) %>%
  slice(chull(PCoA1, PCoA2))

# force consistent factor levels across stage
unique(bd_stage_df$group)
names(stage.3.colors)

lvls <- levels(factor(bd_stage_df$group))

bd_stage_df <- bd_stage_df %>% mutate(group = factor(group, levels = lvls))
hulls <- hulls %>% mutate(group = factor(group, levels = lvls))
bd_stage_centroids <- bd_stage_centroids %>% mutate(group = factor(group, levels = lvls))

# plot
bd_stage_plot <- ggplot(bd_stage_df, aes(PCoA1, PCoA2, color = group)) +
  geom_point(alpha = 0.85) +
  geom_polygon(
    data = hulls,
    aes(fill = group),
    alpha = 0.2,
    color = NA,
    show.legend = FALSE
  ) +
  geom_point(
    data = bd_stage_centroids,
    shape = 4
  ) +
  stat_ellipse(level = 0.68, linewidth = 0.8) +
  coord_equal() +
   scale_x_continuous(
    breaks = c(-30, 0, 30)
  ) +
  scale_y_continuous(
    breaks = c(-20, 0, 20)
  ) +
  scale_color_manual(values = stage.3.colors) +
  scale_fill_manual(values = stage.3.colors) +
  theme_sleek_axe() +
  guides(
    color = guide_legend(reverse = TRUE),
    fill = guide_legend(reverse = TRUE)
  )+
  labs(
    title = "Stage",
    x = xlab,
    y = ylab,
    color = "stage"
  )

geom_defaults()
bd_stage_plot
ggsave(
  "../../figs/betadisp_stage.png",
  width=6,
  height=4,
  dpi=600,
  plot=bd_stage_plot
)
```

```{r stage-pptx, fig.width=16, fig.height=9}
geom_pptx()
betadisp_stage_pptx <- bd_stage_plot + theme_pptx()
betadisp_stage_pptx
ggsave(
  "../../figs/betadisp_stage_pptx.png",
  width=16, 
  height=9, 
  dpi=600,
  plot=betadisp_stage_pptx
)
```



### leachate x stage

```{r}
set.seed(02192026)
bd_int <- betadisper(aitchison_dist, group = grp_int, type = "centroid")

# Permutation test with optional blocking
perm_int <- permutest(bd_int, permutations = 999, pairwise = FALSE)
                  #strata = meta_filt$spawn_night)  # optional
perm_int
```

::: callout-note
A permutation test for homogeneity of multivariate dispersions revealed no significant difference in dispersion across the interaction between leachate and stage (*F*~11,51~ = 0.3491, *p* = 0.971, 999 permutations)
:::

```{r}
plot(bd_int, hull = TRUE, ellipse = TRUE, label = FALSE, cex = 0.8)
```

```{r}
# sample scores (vectors)
bd_int_df <- as.data.frame(bd_int$vectors[, 1:2])
bd_int_df$group <- bd_int$group

# centroid scores
bd_int_centroids <- as.data.frame(bd_int$centroids[, 1:2])
bd_int_centroids$group <- factor(rownames(bd_int_centroids), levels = levels(bd_int$group))

# axis labels with %variance explained
eig <- bd_int$eig
pct <- eig / sum(eig) * 100
xlab <- sprintf("PCoA 1 (%.1f%%)", pct[1])
ylab <- sprintf("PCoA 2 (%.1f%%)", pct[2])

# hulls
hulls <- bd_int_df %>%
  group_by(group) %>%
  slice(chull(PCoA1, PCoA2))

# plot
ggplot(bd_int_df, aes(PCoA1, PCoA2, color = group)) +
  geom_point(size = 2.6, alpha = 0.85) +
  geom_polygon(data = hulls, aes(fill = group), alpha = 0.18, color = NA, show.legend = FALSE) +
  geom_point(data = bd_int_centroids, shape = 4, stroke = 1.2, size = 5) +
  stat_ellipse(level = 0.68, linewidth = 0.8) +
  coord_equal() +
  theme_minimal(base_size = 14) +
  labs(title = "betadisper (int)", x = xlab, y = ylab, color = "int")
```

## Perform PERMANOVA

PERMANOVA mainly tests centroid location differences among groups (i.e., group separation).

betadisper tests dispersion (variance) differences among groups.

If PERMANOVA is significant and betadisper is non-significant, the signal likely reflects location (good).

If both are significant, some of the PERMANOVA “difference” could be due to dispersion heterogeneity (groups with different spreads).
Report both, and interpret cautiously.

> In adonis2, the strata argument restricts permutations to occur within levels of the strata variable (here, spawn night).
> This controls for the grouping effect by ensuring samples are only permuted among others within the same spawn night, which accounts for the non-independence of samples collected on the same night when calculating significance of other factors.
> `adonis2` does not currently support specifying random effects in the model formula like mixed models do.
> So you cannot directly include spawn night as a random effect.Using strata is a practical approach to avoid inflating Type I error due to pseudoreplication or clustering of samples within levels of spawn night and tests effects of leachate and stage while controlling for spawn night as a grouping factor.

```{r}
set.seed(02192026)
# Perform PERMANOVA on Aitchison distance
# Test effects of leachate, stage, and their interaction (blocking permutations by spawn night)
permanova_int.stage <- adonis2(
  aitchison_dist ~ leachate * stage,
  data = meta_filt,
  permutations = 999,
  by = "terms",
  strata = meta_filt$spawn_night
)

print(permanova_int.stage)

# Save results
write.csv(as.data.frame(permanova_int.stage), 
          "../../output/permanova/perm_result_int.stage.csv",
          row.names = TRUE)
```

```{r}
set.seed(02192026)
# Additional PERMANOVA treating stage as continuous time 
permanova_int.hpf <- adonis2(
  aitchison_dist ~ leachate * hpf,
  data = meta_filt,
  permutations = 999,
  by = "terms",
  strata = meta_filt$spawn_night # control for parental cross
)

print(permanova_int.hpf)

# Save results
write.csv(as.data.frame(permanova_int.hpf), 
          "../../output/permanova/perm_result_int.hpf.csv",
          row.names = TRUE)
```

Whether we use developmental stage (categorical factor) or hours post fertilization (continuous numerical) the results are the same...
stage/hpf is significant and leachate and the interactions between leachate and stage/hpf are not.

# Summary

This analysis implemented the key recommendations from Gloor et al. 2017 for compositional microbiome data analysis:

1.  **CLR Transformation**: Applied centered log-ratio transformation to convert compositional data to Euclidean space
2.  **Aitchison Distance**: Calculated Euclidean distance on CLR-transformed data as the appropriate dissimilarity metric
3.  **PCA Biplot**: Created variance-based PCA visualizations on CLR-transformed data
4.  **PERMANOVA**: Tested for significant differences in microbial composition across experimental factors


```{r patch-plots, fig.width=5, fig.height=9}
geom_defaults()
bd_plots <- (bd_stage_plot / bd_night_plot / bd_leachate_plot) +
  plot_annotation(tag_levels = "A") &
  theme(
    plot.tag = element_text(color = "grey30", size = 10),
    legend.position = "right",
    panel.spacing = unit(0.2, "lines")
  ) +
  plot_layout(guides = "collect", heights = c(1,1,1))

bd_plots
```
Save combined plot

```{r}
ggsave("../../figs/betadisper_plots.png", plot = bd_plots, width = 5, height = 9, dpi = 900)
```

# PCoA on Aitchison distance (Euclidean distance on CLR-transformed data) with betadisper hulls and ellipses for spawn night, leachate, and stage.

```{r}

# PCoA
ord <- ape::pcoa(aitchison_dist)

scores <- as.data.frame(ord$vectors[,1:2]) %>% 
  rename(PC1 = Axis.1, PC2 = Axis.2) %>%
  mutate(sample_id = rownames(.)) %>%
  left_join(meta_filt, by = "sample_id")

eig <- ord$values$Relative_eig * 100

spawnhulls <- scores %>%
  group_by(spawn_night) %>%
  slice(chull(PC1, PC2))
```

```{r pcoa-spawn}
pcoa_spawn <- ggplot(scores, aes(PC1, PC2, color = spawn_night, fill = spawn_night)) +
  geom_point(size = 4, alpha = 0.7) +
  stat_ellipse(
  aes(fill = spawn_night),
  geom = "polygon",
  level = 0.68,
  alpha = 0.2,
  color = NA, 
  show.legend = FALSE
) +
stat_ellipse(
  level = 0.68,
  linewidth = 0.8, 
  show.legend = FALSE
)+
  scale_color_manual(values = night.colors.alt) +
  scale_fill_manual(values = night.colors.alt) +
  labs(x = paste0("PC1 (", round(eig[1], 2), "%)"), 
       y = paste0("PC2 (", round(eig[2], 2), "%)")) +
  guides(fill = "none") +
  coord_equal() +
  
  labs(
    title = "Spawn Night",
    x = xlab,
    y = ylab,
    color = "spawn night"
  )+
  
  theme_sleek_axe()

pcoa_spawn
```

```{r}
pcoa_leachate <- ggplot(scores, aes(PC1, PC2, color = leachate, fill = leachate)) +
  geom_point(size = 4, alpha = 0.7) +
  stat_ellipse(
  aes(fill = leachate),
  geom = "polygon",
  level = 0.68,
  alpha = 0.2,
  color = NA, 
  show.legend = FALSE
) +
stat_ellipse(
  level = 0.68,
  linewidth = 0.8, 
  show.legend = FALSE
)+
  scale_color_manual(values = leachate.colors) +
  scale_fill_manual(values = leachate.colors) +
  labs(x = paste0("PC1 (", round(eig[1], 2), "%)"), 
       y = paste0("PC2 (", round(eig[2], 2), "%)")) +
  guides(fill = "none") +
  coord_equal() +
  
  labs(
    title = "Leachate",
    x = xlab,
    y = ylab,
    color = "leachate"
  )+
  
  theme_sleek_axe()

pcoa_leachate
```

```{r pcoa-stage}
pcoa_stage <- ggplot(scores, aes(PC1, PC2, color = stage, fill = stage)) +
  geom_point(size = 4, alpha = 0.7) +
  stat_ellipse(
  aes(fill = stage),
  geom = "polygon",
  level = 0.68,
  alpha = 0.2,
  color = NA, 
  show.legend = FALSE
) +
stat_ellipse(
  level = 0.68,
  linewidth = 0.8, 
  show.legend = FALSE
)+
  scale_color_manual(values = stage.3.colors) +
  scale_fill_manual(values = stage.3.colors) +
  labs(x = paste0("PC1 (", round(eig[1], 2), "%)"), 
       y = paste0("PC2 (", round(eig[2], 2), "%)")) +
  guides(fill = "none",
         color = guide_legend(reverse = TRUE)) +
  coord_equal() +
  
  labs(
    title = "Stage",
    x = xlab,
    y = ylab,
    color = "stage"
  )+
  
  theme_sleek_axe()

pcoa_stage
```

PCoA combined plots 
```{r pcoa-combo, fig.width=5, fig.height=9}
pcoa_combined <- (pcoa_spawn / pcoa_leachate / pcoa_stage) +
  plot_annotation(tag_levels = "A", 
                  title = "Principal Coordinates Analysis (PCoA) \nof Aitchison Beta Diversity") &
  theme(
    plot.tag = element_text(color = "grey60", size = 11),
    plot.title = element_text(size = 14, color = "grey40", hjust = 0.5, margin = margin(b = 20)),
    legend.position = "right",
    panel.spacing = unit(0.2, "lines")
  ) +
  
  plot_layout(guides = "collect", heights = c(1,1,1))

pcoa_combined
```

```{r}
ggsave("../../figs/pcoa_combined.png", plot = pcoa_combined, width = 5, height = 9, dpi = 900)
```


```{r save-patch-plot}
ggsave("../../figs/betadisp_patched_plots.png", 
       plot = bd_plots, 
       width=5,
       height = 9,
       dpi = 900)
```

```{r}
geom_defaults()
bd_plots_pptx <- (bd_stage_plot / bd_night_plot / bd_leachate_plot) +
  plot_annotation(tag_levels = "A") &
  theme(
    plot.tag = element_text(color = "grey30", size = 10),
    legend.position = "right",
    panel.spacing = unit(0.2, "lines")
  ) +
  plot_layout(guides = "collect", heights = c(1,1,1))

bd_plots_pptx
```

## Key Findings

-   PCA plots visually show clustering patterns by Developmental stage (aka timepoint/ hours post fertilization), Night of Spawn, and Parental cross (the later two are confounded)
-   betadispersion results indicate significant difference in variance (dispersion) across night of spawn
-   PERMANOVA results (blocked across Night of Spawn) indicate significant effects of Developmental stage (aka timepoint/ hours post fertilization), and non-significance for leachate and leahcate:stage interactions

## Next Steps

-   Differential abundance analysis (MaAslin3)
-   Conduct PERMANOVA on differentially abundant taxa
-   Functional annotation of differentially abundant taxa (picrust)
-   Differential abundance of functions (functional enrichment) (MaAslin3)

## Output Files

All results and visualizations have been saved to the `../output/permanova` directory: - PCA biplots - PERMANOVA results tables

# Session Info

```{r}
sessionInfo()
```
