---
title: "Core Metrics Visualization"
author: "Sarah Tanja"
date: 2025-10-15
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = ""# Prevents appending '##' to beginning of lines in code output
  )        
```

# Background

This document visualizes core diversity metrics from the filtered QIIME2
analysis (`core-metrics-results-filtered/`). The filtered dataset
excludes chloroplast, mitochondria, and unassigned sequences.

Core metrics include: - **Alpha Diversity**: - Shannon diversity -
Faith's Phylogenetic Diversity, - Evenness - Observed Features

-   **Beta Diversity**:
    -   Bray-Curtis
    -   Jaccard
    -   Weighted UniFrac
    -   Unweighted UniFrac

# Setup

## Install Packages

```{r}
#install.packages("devtools")
#devtools::install_github("jbisanz/qiime2R")
#install.packages("tidyverse")
#install.packages("ggplot2")
#install.packages("patchwork")
#install.packages("ggtext")
#install.packages("viridis")
#install.packages("stringr")
#install.packages("broom")
```

## Load libraries

```{r}
#| message: false
#| warning: false
library(qiime2R)
library(ggplot2)
library(tidyverse)
library(patchwork)
library(stringr)
library(ggtext)
library(viridis)
library(broom)
library(ggsidekick)
library(ggbeeswarm)
library(ggsignif)
```

## Set Paths

```{r}
# Base path to core metrics results
core_metrics_path <- "../../salipante/241121_StonyCoral/core-metrics-results-filtered"
metadata_path <- "../../metadata/metadata.csv"

# Output directory for figures
figures_dir <- "../../figs"

# Create figures directory if it doesn't exist
if (!dir.exists(figures_dir)) {
  dir.create(figures_dir, recursive = TRUE)
}
```

## Set custom ggplot theme

```{r}
theme_sleek_axe <- function() {
  theme_sleek() +
    theme(
      panel.border = element_rect(
        color = "grey70",
        fill = NA,
        linewidth = 0.3
      ),
      axis.line.x  = element_line(color = "grey70"),
      axis.line.y  = element_line(color = "grey70"),
      
      plot.title = element_text(
        size  = 10,
        color = "grey40",
        hjust = 0.5,
        margin = margin(b = 20)
      ),
      
      strip.text = element_text(size = 10, color = "grey40"),
      
      plot.subtitle = element_text(
        size   = 10,
        color  = "grey40",
        hjust  = 0.5,
        margin = margin(b = 40)
      ),
      
      axis.title = element_text(size = 10, color = "grey40"),
      axis.text = element_text(size = 10, color = "grey40"),
      legend.title = element_text(size = 10, color = "grey40"),
      legend.text = element_text(size = 10, color = "grey40"),
  
      plot.margin = margin(t = 0.05, r = 2, b = 0.02, l = 2)
    )
}

```

## Set colorschemes

```{r colors}
leachate.colors <- c(control = "#AEF1FF", 
                low = "#BBC7FF",
                mid = "#7D8BFF", 
                high = "#592F7D")

stage.5.colors <- c(egg = "#FFE362",
                    cleavage = "#EBA600", 
                    morula = "#E6AA83",
                    prawnchip = "#D9685B", 
                    earlygastrula = "#A2223C")

stage.3.colors <- c(cleavage = "#EBA600", 
                  prawnchip = "#D9685B", 
                  earlygastrula = "#A2223C")

hpf.3.colors <- c("4" = "#EBA600", 
                  "9" = "#D9685B", 
                  "14" = "#A2223C")

status.colors <- c(typical = "#75C165", 
                   uncertain = "#E3FAA5", 
                   malformed = "#8B0069")

night.colors <- c(July_6th = "#E3FAA5", 
                  July_7th = "#578B21", 
                  July_8th = "#1E2440")

night.colors.alt <- c(July_6th = "#21918C", 
                  July_7th = "#201158", 
                  July_8th = "#1E2440")
```

## Load Metadata

```{r}
metadata <- read_csv(metadata_path)

# set factors
metadata <- metadata %>% 
  mutate(
    stage = stage %>%
      trimws() %>%
      gsub("\\s+", "", .) %>% # remove spaces
      factor(levels = c("cleavage", "prawnchip", "earlygastrula"),
             ordered = TRUE),
    leachate = factor(leachate, levels = c("control","low","mid","high"),
                      ordered = TRUE),
    parents = factor(parents),
    group = factor(group)
    ) %>% 
  rename(SampleID = sample_id)

str(metadata)
```

::: callout-note
-   stage & leachate are categorical
-   hpf & leachate_mgL are numerical
:::

# Alpha Diversity

**Alpha diversity measures the diversity within individual samples.**

## Shannon Diversity Index

Shannon diversity accounts for both abundance and evenness of species
present.

```{r, shannon-data}
# Load Shannon diversity vector
shannon <- read_qza(file.path(core_metrics_path, "shannon_vector.qza"))

# Create data frame
shannon_df <- shannon$data %>%
  rownames_to_column("SampleID") %>%
  rename(Shannon = shannon_entropy) %>%
  left_join(metadata, by = "SampleID")

head(shannon_df)
```

### Statistical tests

#### Check for normality

```{r}
ggplot(shannon_df, aes(x=Shannon))+
  geom_density()+
  theme_sleek_axe()

shapiro.test(shannon_df$Shannon) # normality
```

::: callout-note
**The Shannon diversity values appear approximately normal** and the
**Shapiro–Wilk test for normality returned a p = 0.68** ( we fail to
reject normality), so we can safely use **parametric tests**
:::

#### Run a two-way ANOVA

Tests main effects and interaction

::: callout-note
A note on categorical vs numeric factors

Use **factors for discrete groups**; use numeric when you want to test a
continuous trend (e.g., linear dose–response).
:::

| Variable | Nature | Should be coded as | Why |
|:----------------|:----------------|:----------------|:---------------------|
| **`stage` (4 hpf, 9 hpf, 14 hpf)** | Distinct developmental stages | **Categorical (factor)** | The stages are not on a linear scale — differences are categorical, not continuous. |
| **`leachate` (0, 0.01, 0.1, 1 mg/L)** | Nominally continuous, used as treatment bins | **Categorical (factor)** for ANOVA **or** **numeric** to test a dose–response trend | If you care about *group differences*, treat as a factor; if you want to test whether Shannon changes *linearly/non-linearly* with concentration, treat it as numeric. |

: How to encode variables for alpha-diversity modeling.

::: callout-note
Since we are looking for non-monotonic (non dose-dependent responses) we
choose to code our factors as categorical.
:::

```{r}
levels(shannon_df$leachate)
levels(shannon_df$stage)
```

```{r}
aov_shannon2 <- aov(Shannon ~ leachate * stage, data = shannon_df)
summary(aov_shannon2)
```

::: callout-note
A two-way analysis of variance (ANOVA) revealed no significant effects
of PVC leachate concentration, developmental stage, or their interaction
on Shannon diversity. Differences in Shannon diversity among leachate
treatments were not statistically significant (*F*₃,₄₈ = 1.25, *p* =
0.302), nor were differences among developmental stages (*F*₂,₄₈ = 1.11,
*p* = 0.339). Similarly, the leachate × stage interaction was not
significant (*F*₆,₄₈ = 0.33, *p* = 0.919), indicating that the effect of
leachate on microbial alpha diversity did not differ among stages.

Two-way ANOVA indicated no significant effects of leachate
concentration, developmental stage, or their interaction on Shannon
diversity (*p* \> 0.3 for all factors).

A two-way analysis of variance (ANOVA) was conducted to test the effects
of categorical factors — **PVC leachate concentration** (control, low,
mid, high) and **developmental stage** (cleavage, prawnchip,
earlygastrula) — on Shannon diversity. Both factors were modeled as
fixed, categorical factors.
:::

... but what if they were linear?

```{r}
class(shannon_df$leachate_mgL)
class(shannon_df$hpf)
```

```{r}
aov_shannon2.num <- aov(Shannon ~ leachate_mgL * hpf, data = shannon_df)
summary(aov_shannon2.num)
```

::: callout-note
Treating leachate concentration and developmental stage as categorical
factors yielded no significant main or interaction effects on Shannon
diversity. When the same variables were treated as continuous, results
were qualitatively similar: no significant effects of leachate_mgL
(*F*₁,₅₆ = 1.10, *p* = 0.300), hpf (*F*₁,₅₆ = 1.55, *p* = 0.218), or
their interaction (*F*₁,₅₆ = 0.04, *p* = 0.841). Thus, alpha diversity
remained stable across both categorical treatments and continuous
gradients, indicating limited sensitivity of microbial diversity to
leachate exposure or developmental progression.
:::

### Shannon Boxplots

```{r shan1-leachates}
# Box plot by leachate treatment
shan1 <- ggplot(shannon_df, aes(x = leachate, y = Shannon,
                                fill = leachate, color=leachate)
                ) +
  
  geom_violin(alpha = 0.5, 
              width = 0.9,
              show.legend=FALSE) +
  
  geom_boxplot(color="black", 
               width = 0.2, 
               outlier.shape = NA, 
               show.legend=FALSE) +
  
  scale_color_manual(
    values = leachate.colors
  ) +
  
  scale_fill_manual(
    values = leachate.colors
  ) +
  
  scale_x_discrete(labels = c(
    control = "0",
    low = "0.01",
    mid = "0.1",
    high = "1"
  ))+
  
  labs(
    title = "",
    x = "",
    y = ""
  ) +
  theme_sleek_axe()

shan1
```

```{r shan2-stage}
# Box plot by timepoint
shan2 <- ggplot(shannon_df, aes(x = factor(hpf), y = Shannon, 
                                fill = factor(hpf), color = factor(hpf))) +
  
  geom_violin(alpha = 0.5, 
              width = 0.9,
              show.legend=FALSE) +
  
  geom_boxplot(color="black", 
               width = 0.2, 
               outlier.shape = NA, 
               show.legend=FALSE) +
  
  scale_fill_manual(values = hpf.3.colors) +
  scale_color_manual(values = hpf.3.colors) +
  
  labs(
    title = "",
    x = "",
    y = ""
  ) +
  theme_sleek_axe() 

shan2
```

```{r}
box1 <- 4
box2 <- 3
box3 <- 12
```

```{r shan3-int}
# Box plot by leachate x timepoint interaction

shan3 <- ggplot(shannon_df, aes(
  x = factor(hpf),
  y = Shannon,
  fill = factor(leachate), 
  color = factor(leachate)
  )) +
  
  geom_violin(aes(group = interaction(hpf, leachate)),
              alpha = 0.5, 
              width = 3,
              position=position_dodge(width=0.6),
              show.legend=FALSE) +
  
  geom_boxplot(aes(group = interaction(hpf, leachate)),
               color="black",
               alpha = 1,
               width = 0.2, 
               outlier.shape = NA, 
               position=position_dodge(width=0.6),
               show.legend=FALSE) +
  
   scale_x_discrete(expand = expansion(mult = c(0.02, 0.02))) +
  
  scale_color_manual(values = leachate.colors) +
  scale_fill_manual(values = leachate.colors) +
  labs(
    title = "Shannon",
    x = "",
    y = ""
      )+
  theme_sleek_axe()

shan3
```

```{r shannon-combo, fig.width=10, fig.height=4}
# Combine plots using patchwork
shannon_combo <- shan3 + shan1 + shan2 +
  plot_layout(widths = c(box3, box1, box2)) +
  plot_annotation(
    title = "Shannon Diversity Index",
    tag_levels = "a",
    theme = theme(
      plot.title = element_text(size = 12,
                                color = "grey40", 
                                hjust = 0.5), 
      plot.tag = element_text(size = 10, 
                              color = "grey40", 
                              hjust = 1, 
                              vjust = -4)
    )
  )

shannon_combo
```

```{r save-shannon}
# Save plots
ggsave(filename = file.path(figures_dir, "shannon_by_leachate.png"), plot = shan1, width = 10, height = 6, dpi = 300)
ggsave(filename = file.path(figures_dir, "shannon_by_timepoint.png"), plot = shan2, width = 10, height = 6, dpi = 300)
ggsave(filename = file.path(figures_dir, "shannon_leachate_x_timepoint.png"), plot = shan3, width = 10, height = 6, dpi = 300)
ggsave(filename = file.path(figures_dir, "shannon_combined.png"), plot = shannon_combo, width = 10, height = 4, dpi = 900)
```


## Faith's Phylogenetic Diversity

Faith's PD measures the phylogenetic diversity of species in a sample.

```{r, faith-data}
# Load Faith's PD vector
faith <- read_qza(file.path(core_metrics_path, "faith_pd_vector.qza"))

# Create data frame
faith_df <- faith$data 

# View it
head(faith_df)
```

```{r, faith-metadata}
# Join to metadata
faith_df <- faith_df %>%
  rename("SampleID"=V1) %>%
  rename(Faith=V2) %>%
  left_join(metadata, by ="SampleID")
```

## Statistical tests

### Check for normality

```{r}
ggplot(faith_df, aes(x=Faith))+
  geom_density()+
  theme_sleek_axe()

shapiro.test(faith_df$Faith) # normality
```

::: callout-note
Check normality of Faith's Phylogenetic Diversity values using
Shapiro-Wilk test. Since p = 0.78, we fail to reject the null hypothesis
of normality, suggesting that Faith's PD values are approximately
normally distributed and suitable for parametric testing.
:::

### Run a two-way ANOVA

#### Using categorical variables

```{r}
aov_faith2 <- aov(Faith ~ leachate * stage, data = faith_df)
summary(aov_faith2)
```

::: callout-note
Two-way ANOVA with Faith's Phylogenetic Diversity as the response
variable and categorical factors (PVC leachate level and developmental
stage) show that leachate level did not effect Faith's PD (*F*₃,₄₈ =
1.31, *p* = 0.280), while developmental stage *did* (*F*₂,₄₈ = 3.63, *p*
= 0.034). However, the interaction between leachate and stage was not
significant (*F*₆,₄₈ = 0.238, *p* = 0.962), indicating that the effect
of leachate on Faith's PD did not differ across developmental stages.
:::

#### Using numerical variables

```{r}
aov_faith2.num <- aov(Faith ~ leachate_mgL * hpf, data = faith_df)
summary(aov_faith2.num)
```

::: callout-note
Two-way ANOVA with Faith's Phylogenetic Diversity treating timepoint as
a continuous variable showed similar results as categorical factors.
Hours post-fertilization did have an impact on within-sample
phylogenetic diversity (*F*~1,56~ = 6.72, *p* = 0.01), while leachate
(*F~1,56~* = 0.68, *p* = 0.41) and the interaction between leachate and
time (*F*~1,56~ = 0.3, *p* = 0.58) did not.
:::

Since stage was significant we can do a Tukey HSD to see which stages
separate from each other

```{r}
TukeyHSD(aov_faith2)
```

::: callout-important
It appears the only sig difference between groups is between the
cleavage and early gastrula stages (Tukey HSD *padj* = 0.04)
:::

### Faith Boxplots

```{r fait1-leachate}
# Box plot by leachate treatment
fait1 <- ggplot(faith_df, aes(
  x = factor(leachate),
  y = Faith,
  fill = factor(leachate), 
  color=factor(leachate))
  ) +

  geom_violin(alpha = 0.5, 
              width = 0.9,
              show.legend=FALSE) +
  
  geom_boxplot(color="black", 
               width = 0.2, 
               outlier.shape = NA, 
               show.legend=FALSE) +
  
    
  scale_fill_manual(values=leachate.colors) +
  scale_color_manual(values=leachate.colors)+
  
  scale_x_discrete(labels = c(
    control = "0",
    low = "0.01",
    mid = "0.1",
    high = "1"
  ))+
  
  scale_y_continuous(breaks = seq(0, 170, by = 50), 
                     limits = c(55, 160)) +
  
  labs(title = "",
       x = "",
       y = ""
       )+
  theme_sleek_axe()
  
fait1  
```

```{r fait2-stage}
# Box plot by hpf
fait2 <- ggplot(faith_df, aes(x = factor(hpf), 
                              y = Faith, 
                              fill = factor(hpf), 
                              color=factor(hpf))) +

  geom_violin(alpha = 0.5, 
              width = 1,
              show.legend=FALSE) +
  
  geom_boxplot(color="black", 
               width = 0.2, 
               outlier.shape = NA, 
               show.legend=FALSE) +
  
  geom_signif(
    comparisons = list(c("4", "14")),
    annotations = "*",
    y_position = 150,
    tip_length = 0.02,
    textsize = 6, 
    color = "grey20",
    show.legend = FALSE
  ) +
  
  scale_fill_manual(values = hpf.3.colors) +
  scale_color_manual(values = hpf.3.colors)+
  
  scale_y_continuous(breaks = seq(0, 170, by = 50), 
                     limits = c(55, 160)) +
  
  labs(
    title = "",
    x = "",
    y = ""
  ) +
  theme_sleek_axe()

fait2
```

```{r fait3-int}
# Box plot by leachate x timepoint interaction

fait3 <- ggplot(faith_df, aes(
  x = factor(hpf),
  y = Faith,
  fill = factor(leachate),
  color = factor(leachate)
)) +

  geom_violin(aes(group = interaction(hpf, leachate)),
              alpha = 0.5, 
              width = 3,
              position=position_dodge(width=0.6),
              show.legend=FALSE) +
  
  geom_boxplot(aes(group = interaction(hpf, leachate)),
               color="black",
               alpha = 1,
               width = 0.2, 
               outlier.shape = NA, 
               position=position_dodge(width=0.6),
               show.legend=FALSE) +
  
  scale_x_discrete(expand = expansion(mult = c(0.02, 0.02))) +
  scale_y_continuous(breaks = seq(0, 170, by = 50), 
                     limits = c(55, 160)) +
  
  scale_fill_manual(values = leachate.colors) +
  scale_color_manual(values = leachate.colors) +
  labs(
    title = "Faith",
    x = "",
    y = ""
  ) +
  theme_sleek_axe()

fait3
```

```{r fait-combo, fig.width=10, fig.height=4}
# Combine plots using patchwork
faith_combo <- fait3 + fait1 + fait2+
  plot_layout(widths = c(box3, box1, box2)) +
  plot_annotation(
    title = "Faith's Phylogenetic Diversity",
    theme = theme(
      plot.title = element_text(size = 12,
                                color = "grey40", 
                                hjust = 0.5)
    )
  )

faith_combo
```

```{r}
# Save plots
ggsave(filename = file.path(figures_dir, "faith_by_leachate.png"), plot = fait1, width = 10, height = 6, dpi = 900)
ggsave(filename = file.path(figures_dir, "faith_by_timepoint.png"), plot = fait2, width = 10, height = 6, dpi = 900)
ggsave(filename = file.path(figures_dir, "faith_leachate_x_timepoint.png"), plot = fait3, width = 10, height = 6, dpi = 900)
ggsave(filename = file.path(figures_dir, "faith_combined.png"), plot = faith_combo, width = 10, height = 4, dpi = 900)
```

## Pielou's Evenness

Evenness measures how equal species abundances are within a sample.

```{r, pielou-data}
# Load evenness vector
evenness <- read_qza(file.path(core_metrics_path, "evenness_vector.qza"))

# Create data frame
evenness_df <- evenness$data 

# View it
head(evenness_df)
```

```{r}
evenness_df <- evenness_df %>% 
  rownames_to_column("SampleID") %>%
  rename(Evenness = pielou_evenness) %>%
  left_join(metadata, by = "SampleID")

head(evenness_df)
```

## Statistical tests

### Check for normality

```{r}
ggplot(evenness_df, aes(x=Evenness))+
  geom_density()+
  theme_sleek_axe()
shapiro.test(evenness_df$Evenness) # normality
```

::: callout-note
Check normality of Pielou's Evenness values using Shapiro-Wilk test.
Since *p* = 0.09, we fail to reject the null and conclude that Pielou's
Evenness values are approximately normally distributed, allowing for
parametric testing.
:::

### Run a two-way ANOVA

#### Using categorical variables

```{r}
aov_evenness2 <- aov(Evenness ~ leachate * stage, data = evenness_df)
summary(aov_evenness2)
```

::: callout-note
Two-way ANOVA with Pielou's Evenness as the response variable and
categorical factors (PVC leachate level and developmental stage)
revealed no differences in evenness (equal species abundances) within
samples across leachate treatments (*F*₃,₄₈ = 0.75, *p* = 0.53),
developmental stages (*F*₂,₄₈ = 1.02, *p* = 0.37), or their interaction
(*F*₆,₄₈ = 0.29, *p* = 0.94). This suggests that species evenness
remained stable regardless of leachate exposure *or* developmental
progression from 4 to 14 hours post-fertilization.
:::

#### Using numerical variables

```{r}
aov_evenness2.num <- aov(Evenness ~ leachate_mgL * hpf, data = evenness_df)
summary(aov_evenness2.num)
```

::: callout-note
Two-way ANOVA with Pielou's Evenness treating leachate concentration and
timepoint as continuous variables showed similar results as categorical
treatment of explanatory factors. All *p* \> 0.1, indicating no
significant effects of leachate concentration, developmental timepoint,
or their interaction on species evenness within samples.
:::

### Boxplots

```{r piel1-leachate}
# Box plot by leachate
piel1 <- ggplot(evenness_df, aes(x = factor(leachate), y = Evenness, fill = factor(leachate), color=factor(leachate))) +
  
  geom_violin(alpha = 0.5, 
              width = 0.9,
              show.legend=FALSE) +
  
  geom_boxplot(color="black", 
               width = 0.2, 
               outlier.shape = NA, 
               show.legend=FALSE) +
  
  scale_fill_manual(values = leachate.colors) +
  scale_color_manual(values = leachate.colors)+
  
  labs(
    title = "",
    x = "",
    y = ""
  ) +
  
  scale_x_discrete(labels = c(
    control = "0",
    low = "0.01",
    mid = "0.1",
    high = "1"
  ))+
  
  theme_sleek_axe()

piel1
```

```{r piel2-stage}
# Box plot by stage
piel2 <- ggplot(evenness_df, aes(x = factor(hpf), y = Evenness, 
                                 fill = factor(hpf), color=factor(hpf))
                )+
 
  geom_violin(alpha = 0.5, 
              width = 0.9,
              show.legend=FALSE) +
  
  geom_boxplot(color="black", 
               width = 0.2, 
               outlier.shape = NA, 
               show.legend=FALSE) +
  
  scale_fill_manual(values = hpf.3.colors) +
  scale_color_manual(values = hpf.3.colors) +
  
  labs(
    title = "",
    x = "",
    y = ""
  ) +
  theme_sleek_axe()

piel2
```

```{r piel3-int}
# Box plot by leachate x timepoint interaction
piel3 <- ggplot(evenness_df, aes(x = factor(hpf), y = Evenness, fill = leachate, color=leachate)) +
 
  geom_violin(aes(group = interaction(hpf, leachate)),
              alpha = 0.5, 
              width = 1.5,
              position=position_dodge(width=0.6),
              show.legend=FALSE) +
  
  geom_boxplot(aes(group = interaction(hpf, leachate)),
               color="black",
               alpha = 1,
               width = 0.2, 
               outlier.shape = NA, 
               position=position_dodge(width=0.6),
               show.legend=FALSE) +
  
  scale_color_manual(values = leachate.colors) +
  scale_fill_manual(values = leachate.colors) +
  
  labs(
    title = "Pielou's",
    x = "",
    y = ""
  ) +
  theme_sleek_axe()

piel3
```

```{r piel-combo, fig.width=10, fig.height=4}
# Combine plots using patchwork
pielou_combo <- piel3+ piel1 + piel2+
  plot_layout(widths = c(box3, box1, box2)) +
  plot_annotation(
    title = "Pielou's Evenness",
    theme = theme(
      plot.title = element_text(size = 12,
                                color = "grey40", 
                                hjust = 0.5)
    )
  )
pielou_combo
```

```{r}
# Save plots
ggsave(filename = file.path(figures_dir, "pielou_by_leachate.png"), plot = piel1, width = 10, height = 6, dpi = 900)
ggsave(filename = file.path(figures_dir, "pielou_by_timepoint.png"), plot = piel2, width = 10, height = 6, dpi = 900)
ggsave(filename = file.path(figures_dir, "pielou_leachate_x_timepoint.png"), plot = piel3, width = 10, height = 6, dpi = 900)
ggsave(filename = file.path(figures_dir, "pielou_combined.png"), plot = pielou_combo, width = 10, height = , dpi = 900)
```

## Observed Features (ASV Richness)

The number of unique amplicon sequence variants (ASVs) observed in each
sample.

```{r, asv-rich-data}
# Load observed features vector
observed <- read_qza(file.path(core_metrics_path, "observed_features_vector.qza"))

# Create data frame
observed_df <- observed$data 

# View it
head(observed_df)
```

```{r, asv-rich-metadata}
observed_df <- observed_df %>%
  rownames_to_column("SampleID") %>%
  rename(ObservedFeatures = observed_features) %>%
  left_join(metadata, by = "SampleID")
```

## Statistical tests

### Check for normality

```{r}
ggplot(observed_df, aes(x=ObservedFeatures))+
  geom_density()+
  theme_sleek_axe()
shapiro.test(observed_df$ObservedFeatures) # normality
```

::: callout-note
Check normality of Observed Features values using Shapiro-Wilk test..
since *p* = 0.12, we can reject the null hypothesis and conclude that
Observed Features values are approximately normally distributed,
allowing for parametric testing.
:::

### Run a two-way ANOVA

#### Using categorical variables

```{r}
aov_observed2 <- aov(ObservedFeatures ~ leachate * stage, data = observed_df)
summary(aov_observed2)
```

::: callout-note
Two-way ANOVA with Observed Features as the response variable and
categorical factors (PVC leachate level and developmental stage) show no
differences between groups (all *p* \> 0.15)
:::

#### Using numerical variables

```{r}
aov_observed2.num <- aov(ObservedFeatures ~ leachate_mgL * hpf, data = observed_df)
summary(aov_observed2.num)
```

::: callout-note
Two-way ANOVA with Observed Features treating leachate concentration and
timepoint as continuous variables show similar results, all *p* \> 0.11
:::

### Boxplots

```{r asv1-leachate}
# Box plot by leachate treatment
asv1 <- ggplot(observed_df, 
               aes(x = leachate, y = ObservedFeatures, 
                   fill = leachate, color=leachate)) +
  
  geom_violin(alpha = 0.5, 
              width = 0.9, 
              outlier.shape = NA, 
              show.legend=FALSE) +
  
  geom_boxplot(color="black", 
               width = 0.2, 
               outlier.shape = NA, 
               show.legend=FALSE) +
  
  scale_color_manual(values = leachate.colors) +
  scale_fill_manual(values = leachate.colors) +
  
  scale_x_discrete(labels = c(
    control = "0",
    low = "0.01",
    mid = "0.1",
    high = "1"
  ))+
  
  labs(
    title = "",
    x = "",
    y = ""
  ) +
  theme_sleek_axe()

asv1
```

```{r asv2-stage}
# Box plot by timepoint
asv2 <- ggplot(observed_df, aes(x = factor(hpf), y = ObservedFeatures, 
                                fill = factor(hpf), color=factor(hpf))) +
  
  geom_violin(
              alpha = 0.5, 
              width = 0.9, 
              outlier.shape = NA, 
              show.legend=FALSE) +
  
  geom_boxplot(
               color="black",
               alpha=1,
               width = 0.2, 
               outlier.shape = NA, 
               show.legend=FALSE) +
  
  scale_fill_manual(values = hpf.3.colors) +
  scale_color_manual(values = hpf.3.colors)+
  
  labs(
    title = "",
    x = "",
    y = ""
  ) +
  theme_sleek_axe()

asv2
```

```{r asv-int-asv3}
# Box plot by leachate x timepoint interaction
asv3 <- ggplot(observed_df, aes(x = factor(hpf), y = ObservedFeatures,
                                fill = leachate, color=leachate)) +
  
  geom_violin(aes(group = interaction(hpf, leachate)),
              alpha = 0.5, 
              width = 1.5,
              position=position_dodge(width=0.8),
              show.legend=FALSE) +
  
  geom_boxplot(aes(group = interaction(hpf, leachate)),
               color="black",
               alpha = 1,
               width = 0.2, 
               outlier.shape = NA, 
               position=position_dodge(width=0.8),
               show.legend=FALSE) +
  
   scale_x_discrete(expand = expansion(mult = c(0.02, 0.02))) +
  
  #geom_beeswarm(aes(group = interaction(hpf, leachate)),
  #            dodge.width = 0.7, 
  #            size = 1.5, 
  #            alpha = 1, 
  #            show.legend = FALSE) +

  scale_color_manual(values = leachate.colors) +
  scale_fill_manual(values = leachate.colors)+
    
  labs(
    title = "",
    x = "",
    y = ""
      ) +
  theme_sleek_axe()

asv3  
```

```{r asv-combo, fig.width=10, fig.height=4}
asv_combo <- asv3 + asv1 + asv2 +
  plot_layout(widths = c(box3, box1, box2))+
  plot_annotation(
    title = "Observed Features (ASV Richnness)",
    theme = theme(
      plot.title = element_text(size = 14, color = "grey40",  hjust = 0.5)
    )
  )
asv_combo
```


```{r}
# Save plots
ggsave(filename = file.path(figures_dir, "observed_features_by_leachate.png"), plot = asv1, width = 10, height = 6, dpi = 900)
ggsave(filename = file.path(figures_dir, "observed_features_by_timepoint.png"), plot = asv2, width = 10, height = 6, dpi = 900)
ggsave(filename = file.path(figures_dir, "observed_features_leachate_x_timepoint.png"), plot = asv3, width = 10, height = 6, dpi = 900)
ggsave(filename = file.path(figures_dir, "observed_features_combined.png"), plot = asv_combo, width = 10, height = 4, dpi = 900)
```

# Alpha Diversity Summary Panel

## Shannon + Faith
```{r}
alpha_shannon_faith <- shannon_combo / faith_combo +
  plot_annotation(
    title = "Alpha Diversity",
    theme = theme(
      plot.title = element_text(size = 14, color = "grey40", hjust = 0.5)
    )
  )

alpha_shannon_faith
```
```{r}
ggsave(filename = file.path(figures_dir, "alpha_shannon_faith.png"), plot = alpha_shannon_faith, width = 10, height = 6, dpi = 900)
```


## Shannon, Faith, Pielou, Observed Features

```{r, fig.height=8, fig.width=9}
alpha_combined_v2 <- wrap_plots(shannon_combo, faith_combo,
                                pielou_combo, asv_combo, 
                                ncol=1, 
                                nrow=4, 
                                byrow=TRUE) +
  plot_annotation(
    title = "Alpha Diversity Metrics by Leachate Treatment, Timepoint, and Interaction",
    theme = theme(
      plot.title = element_text(size = 16, color = "grey30", hjust = 0.5)
    )
  )

alpha_combined_v2
```

# Beta Diversity

**Beta diversity measures the dissimilarity in community composition
between samples.**



## Bray-Curtis Dissimilarity

Bray-Curtis dissimilarity is based on abundance differences and does not
account for phylogeny.

```{r, braycurtis-data}
# Load Bray-Curtis PCoA results
bc_pcoa <- read_qza(file.path(core_metrics_path, "bray_curtis_pcoa_results.qza"))

# Extract principle component coordinates
bc_coords <- bc_pcoa$data$Vectors
```

```{r, braycurtis-metadata}
bc_coords <- bc_coords %>% 
  left_join(metadata, by = "SampleID")

bc_variance <- bc_pcoa$data$ProportionExplained
```

```{r, braycurtis-PCoAplot}
#| fig-width: 10
#| fig-height: 8

# PCoA plot colored by leachate
bray1 <- ggplot(bc_coords, aes(x = PC1, y = PC2, color = factor(leachate), shape = factor(hpf))) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values=leachate.colors) +
  labs(
    title = "Bray-Curtis PCoA",
    subtitle = "Community composition dissimilarity",
    x = paste0("PC1 (", round(bc_variance[1] * 100, 2), "%)"),
    y = paste0("PC2 (", round(bc_variance[2] * 100, 2), "%)"),
    color = "Leachate",
    shape = "Timepoint"
  ) +
  theme_sleek_axe() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    axis.title = element_text(size = 12),
    legend.position = "right"
  )

# PCoA plot colored by timepoint
bray2 <- ggplot(bc_coords, aes(x = PC1, y = PC2, color = factor(hpf), shape = factor(leachate)), fill = factor(cross)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values=hpf.3.colors) +
  labs(
    title = "Bray-Curtis PCoA",
    subtitle = "Community composition dissimilarity",
    x = paste0("PC1 (", round(bc_variance[1] * 100, 2), "%)"),
    y = paste0("PC2 (", round(bc_variance[2] * 100, 2), "%)"),
    color = "Timepoint (hpf)",
    shape = "Leachate"
  ) +
  theme_sleek_axe() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    axis.title = element_text(size = 12),
    legend.position = "right"
  )

bray1
bray2
```

```{r}
# Save plots
ggsave(filename = file.path(figures_dir, "bray_curtis_by_leachate.png"), plot = bray1, width = 10, height = 8, dpi = 300)
ggsave(filename = file.path(figures_dir, "bray_curtis_by_timepoint.png"), plot = bray2, width = 10, height = 8, dpi = 300)
```

::: callout-note
44.35% of variation explained by... parental cross? How do we add that
info into the figure?
:::

## Jaccard Distance

Jaccard distance is based on presence/absence and does not account for
phylogeny.

```{r}
# Load Jaccard PCoA results
jac_pcoa <- read_qza(file.path(core_metrics_path, "jaccard_pcoa_results.qza"))

# Extract coordinates and variance explained
jac_coords <- jac_pcoa$data$Vectors 

# View 
head(jac_coords)
```

```{r}
jac_coords <- jac_coords %>%
  left_join(metadata, by = "SampleID")

head(jac_coords)

jac_variance <- jac_pcoa$data$ProportionExplained

head(jac_variance)
```

```{r}
#| fig-width: 13
#| fig-height: 8

# PCoA plot colored by leachate
jacc1 <- ggplot(jac_coords, aes(x = PC1, y = PC2, color = factor(leachate), shape = factor(hpf))) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values=leachate.colors) +
  labs(
    title = "Jaccard PCoA",
    subtitle = "Presence/absence dissimilarity",
    x = paste0("PC1 (", round(jac_variance[1] * 100, 2), "%)"),
    y = paste0("PC2 (", round(jac_variance[2] * 100, 2), "%)"),
    color = "Leachate",
    shape = "Timepoint"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    axis.title = element_text(size = 12),
    legend.position = "right"
  )

# PCoA plot colored by timepoint
jacc2 <- ggplot(jac_coords, aes(x = PC1, y = PC2, color = factor(hpf), shape = factor(leachate))) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values=hpf.3.colors) +
  labs(
    title = "Jaccard PCoA",
    subtitle = "Presence/absence dissimilarity",
    x = paste0("PC1 (", round(jac_variance[1] * 100, 2), "%)"),
    y = paste0("PC2 (", round(jac_variance[2] * 100, 2), "%)"),
    color = "Timepoint (hpf)",
    shape = "Leachate"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    axis.title = element_text(size = 12),
    legend.position = "right"
  )

# Save plots
ggsave(filename = file.path(figures_dir, "jaccard_by_leachate.png"), plot = jacc1, width = 10, height = 8, dpi = 300)
ggsave(filename = file.path(figures_dir, "jaccard_by_timepoint.png"), plot = jacc2, width = 10, height = 8, dpi = 300)

jacc1
jacc2
```

## Weighted UniFrac Distance

Weighted UniFrac considers phylogenetic distances and abundance
differences.

```{r}
# Load Weighted UniFrac PCoA results
wuf_pcoa <- read_qza(file.path(core_metrics_path, "weighted_unifrac_pcoa_results.qza"))

# Extract coordinates and variance explained
wuf_coords <- wuf_pcoa$data$Vectors 

head(wuf_coords)
```

```{r}
wuf_coords_n_meta <- wuf_coords %>%
  left_join(metadata, by = "SampleID")

wuf_variance <- wuf_pcoa$data$ProportionExplained
```

```{r}
#| fig-width: 13
#| fig-height: 8

# PCoA plot colored by leachate
wufr1 <- ggplot(wuf_coords_n_meta, aes(x = PC1, y = PC2, color = factor(leachate), shape = factor(hpf))) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values=leachate.colors) +
  labs(
    title = "Weighted UniFrac PCoA",
    subtitle = "Phylogenetic dissimilarity (abundance-weighted)",
    x = paste0("PC1 (", round(wuf_variance[1] * 100, 2), "%)"),
    y = paste0("PC2 (", round(wuf_variance[2] * 100, 2), "%)"),
    color = "Leachate",
    shape = "Timepoint"
  ) +
  theme_sleek_axe() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    axis.title = element_text(size = 12),
    legend.position = "right"
  )

# PCoA plot colored by timepoint
wufr2 <- ggplot(wuf_coords_n_meta, aes(x = PC1, y = PC2, color = factor(hpf), shape = factor(leachate))) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values=hpf.3.colors) +
  labs(
    title = "Weighted UniFrac PCoA",
    subtitle = "Phylogenetic dissimilarity (abundance-weighted)",
    x = paste0("PC1 (", round(wuf_variance[1] * 100, 2), "%)"),
    y = paste0("PC2 (", round(wuf_variance[2] * 100, 2), "%)"),
    color = "Timepoint (hpf)",
    shape = "Leachate"
  ) +
  theme_sleek_axe() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    axis.title = element_text(size = 12),
    legend.position = "right"
  )

# Save plots
ggsave(filename = file.path(figures_dir, "weighted_unifrac_by_leachate.png"), plot = wufr1, width = 10, height = 8, dpi = 900)
ggsave(filename = file.path(figures_dir, "weighted_unifrac_by_timepoint.png"), plot = wufr2, width = 10, height = 8, dpi = 900)

wufr1
wufr2
```

## Unweighted UniFrac Distance

Unweighted UniFrac considers phylogenetic distances but not abundance
(presence/absence only).

```{r}
# Load Unweighted UniFrac PCoA results
uwuf_pcoa <- read_qza(file.path(core_metrics_path, "unweighted_unifrac_pcoa_results.qza"))

# Extract coordinates and variance explained
uwuf_coords <- uwuf_pcoa$data$Vectors %>%
  left_join(metadata, by = "SampleID")

uwuf_variance <- uwuf_pcoa$data$ProportionExplained
```

```{r}
#| fig-width: 10
#| fig-height: 8

# PCoA plot colored by leachate
uwuf1 <- ggplot(uwuf_coords, aes(x = PC1, y = PC2, color = factor(leachate), shape = factor(timepoint))) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_viridis_d(option = "plasma", begin = 0.2, end = 0.9) +
  labs(
    title = "Unweighted UniFrac PCoA",
    subtitle = "Phylogenetic dissimilarity (presence/absence)",
    x = paste0("PC1 (", round(uwuf_variance[1] * 100, 2), "%)"),
    y = paste0("PC2 (", round(uwuf_variance[2] * 100, 2), "%)"),
    color = "Leachate",
    shape = "Timepoint"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    axis.title = element_text(size = 12),
    legend.position = "right"
  )

# PCoA plot colored by timepoint
uwuf2 <- ggplot(uwuf_coords, aes(x = PC1, y = PC2, color = factor(timepoint), shape = factor(leachate))) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_viridis_d(option = "mako", begin = 0.2, end = 0.9) +
  labs(
    title = "Unweighted UniFrac PCoA",
    subtitle = "Phylogenetic dissimilarity (presence/absence)",
    x = paste0("PC1 (", round(uwuf_variance[1] * 100, 2), "%)"),
    y = paste0("PC2 (", round(uwuf_variance[2] * 100, 2), "%)"),
    color = "Timepoint (hpf)",
    shape = "Leachate"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    axis.title = element_text(size = 12),
    legend.position = "right"
  )

# Save plots
ggsave(filename = file.path(figures_dir, "unweighted_unifrac_by_leachate.png"), plot = uwuf1, width = 10, height = 8, dpi = 300)
ggsave(filename = file.path(figures_dir, "unweighted_unifrac_by_timepoint.png"), plot = uwuf2, width = 10, height = 8, dpi = 300)

uwuf1
uwuf2
```

## Beta Diversity Summary Panel

```{r}
#| fig-width: 14
#| fig-height: 12

# Combine all beta diversity PCoA plots by leachate
beta_leachate <- (bray1 + jacc1) / (wufr1 + uwuf1) +
  plot_annotation(
    title = "Beta Diversity PCoA Plots - Colored by Leachate Treatment",
    theme = theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5))
  )

# Combine all beta diversity PCoA plots by timepoint
beta_timepoint <- (bray2 + jacc2) / (wufr2 + uwuf2) +
  plot_annotation(
    title = "Beta Diversity PCoA Plots - Colored by Developmental Timepoint",
    theme = theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5))
  )

# Save plots
ggsave(filename = file.path(figures_dir, "beta_diversity_panel_by_leachate.png"), plot = beta_leachate, width = 14, height = 12, dpi = 300)
ggsave(filename = file.path(figures_dir, "beta_diversity_panel_by_timepoint.png"), plot = beta_timepoint, width = 14, height = 12, dpi = 300)

beta_leachate
beta_timepoint
```

# Summary

This was for initial exploration of alpha and beta diversity metrics in the coral microbiome dataset. We generated a comprehensive set of visualizations to assess how microbial diversity changes with PVC leachate exposure and developmental progression from 4 to 14 hours post-fertilization.

**Alpha Diversity Metrics:** - Shannon Diversity Index - Faith's
Phylogenetic Diversity - Pielou's Evenness - Observed ASV Richness -
Alpha Diversity Summary Panel (all metrics combined)

**Beta Diversity Metrics:** 
- Bray-Curtis dissimilarity
(abundance-based) - Jaccard distance (presence/absence) - Weighted
UniFrac distance (phylogenetic + abundance) - Unweighted UniFrac
distance (phylogenetic presence/absence) - Beta Diversity Summary Panel
(all PCoA plots combined)

Each metric is visualized by both leachate treatment level and
developmental timepoint (hours post fertilization), allowing for
comprehensive assessment of how microbiome diversity changes with PVC
leachate exposure and coral development.

**All plots are saved to the `figs` directory**
