---
title: "Running PICRUSt2"
subtitle: "by creating a picrust2 environment"
author: "Sarah Tanja"
date: 2026-05-03
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = FALSE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "") # Prevents appending '##' to beginning of lines in code output)        
```

# Background

# Load libraries

```{r, eval=TRUE}
library(reticulate)
library(tidyverse)
library(qiime2R)
```



Use qiime tools (qiime2R) to  format feature table as a BIOM and representative sequences as fasta


## Feature table (BIOM)

located at path `../../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_featureTable_filtered.qza`

```{r, engine='bash', eval=TRUE}
qiime tools peek ../../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_featureTable_filtered.qza
```

```{r, engine='bash'}
qiime tools export \
  --input-path ../../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_featureTable_filtered.qza \
  --output-path ../../output/picrust2/
```
::: callout-important
Now the feature table is `feature-table.biom`
:::

## Sequence file

located at path `../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_representative-sequences.qza`

```{r, engine='bash', eval=TRUE}
qiime tools peek ../../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_representative-sequences.qza
```

```{r, engine='bash'}
qiime tools export \
  --input-path ../../salipante/241121_StonyCoral/270x200/250414_StonyCoral_270x200_representative-sequences.qza \
  --output-path ../../output/picrust2/
```
::: callout-important
Now the sequences are `dna-sequences.fasta`
:::

# Run picrust2 on all ASVs full-pipeline
The output of this will get passed through MaAsLin3 again to find differentially abundant or prevalent pathways across our experimental factors
```{r, engine='bash'}
conda activate picrust2 
picrust2_pipeline.py \
   -i ../../output/picrust2/feature-table.biom \
   -s ../../output/picrust2/dna-sequences.fasta \
   -o ../../output/picrust2/envrun \
   -p 4 \
   --verbose

```
