---
title: "Maaslin3 Differential abundance and prevalence testing on PICRUSt2 full functional profile output of predicted MetaCyc metabolic pathways"
author: "Sarah Tanja"
date: 2026-04-30
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = FALSE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "") # Prevents appending '##' to beginning of lines in code output)        
```

# Background

> Predicted pathway abundances were inferred from ASV sequences using
> PICRUSt2 and compared among taxa associated with PVC leachate exposure
> You are modeling predicted function from taxonomy, so you are not
> directly measuring function but rather inferring it based on the
> taxonomy of the microbes present. This is a common approach in
> microbiome studies when direct functional measurements (like
> metagenomics or metatranscriptomics) are not available. “Predicted
> microbial functional potential shifted…”

note to self.. can't use qiime2R on windows so go to raven but can't
update raven r version (needed for maaslin) so head back to minerva to
run maaslin3 \# Load libraries

```{r, eval=TRUE}
library(tidyverse)
library(qiime2R)
library(maaslin3)
```

# Set paths

```{r}
metadata_path <- "../../metadata/meta.csv"
output_path <- "../../output/picrust2"
fig_path <- "../../output/figs"
```

# Load pathway abundance table

```{r}
path <- read_tsv("../../output/picrust2/envrun/pathways_out/path_abun_unstrat.tsv.gz")
```

Make pathway names rownames and remove the pathway column

```{r}
path <- path %>% 
  column_to_rownames(var = "pathway") %>% 
  as.data.frame()
```

Transpose so rownames are samples and columns are pathways

```{r}
path <- t(path)
dim(path)
```

We have 63 samples and 547 pathways to run differential
abundance/prevalence testing on in MaAsLin3. Our 63 samples must match
our metadata!

## Load metadata

```{r}
# Load metadata
metadata <- read_csv(metadata_path)

# set factors
metadata <- metadata %>% 
  mutate(
    collection_date = as.Date(collection_date, format = "%d-%b-%Y"),
    stage    = factor(stage,    levels = c("cleavage", "prawnchip", "earlygastrula"), ordered = TRUE),
    leachate = factor(leachate, levels = c("control", "low", "mid", "high"),        ordered = TRUE),
    spawn_night = factor(
      collection_date,
      levels  = as.Date(c("06-Jul-2024", "07-Jul-2024", "08-Jul-2024"), format = "%d-%b-%Y"),
      labels  = c("July 6th", "July 7th", "July 8th"),
      ordered = TRUE
    )
  )
```

```{r}
# make samples rownames
meta <- metadata %>% 
  column_to_rownames(var = "sample_id") %>% 
  as.data.frame()

# View metadata structure
str(meta)
```

## Run MaAsLin3

```{r maaslin-full}
#set.seed(05032026)
#mp <- maaslin3(
#  input_data = path,
#  input_metadata = meta,
#  output = file.path(output_path, "maaslin_metacycpaths"),
#  formula = ~ leachate*hpf + (1|spawn_night),
#  verbosity = 'ERROR',
#  cores = 3, 
#  plot_associations = FALSE, 
#  save_models = TRUE
#)
```

```{r sig-paths}
sig_paths <- read_tsv(file.path(output_path, "maaslin_metacycpaths/significant_results.tsv"))
```

.L Does the response rise steadily? .Q Does it peak in middle doses? .C
Does it show a more complex non-monotonic pattern?

leachate

```{r sig-leachate}
sig_leachate <- sig_paths %>% 
  filter(metadata == "leachate") %>% 
  arrange(qval_joint)

length(unique(sig_leachate$feature))
unique(sig_leachate$feature) 
```

```{r}
prev_leachate <- sig_leachate %>% 
  filter(model == "prevalence")

length(unique(prev_leachate$feature))

abun_leachate <- sig_leachate %>% 
  filter(model == "abundance")

length(unique(abun_leachate))
```

```{r}
paste0(
  "Features significant for prevalence but not abundance: ",
  paste(
    setdiff(prev_leachate$feature,
            abun_leachate$feature),
    collapse = ", "
  )
)

paste0(
  "Features significant for abundance but not prevalence: ",
  paste(
    setdiff(abun_leachate$feature,
            prev_leachate$feature),
    collapse = ", "
  )
)

paste0(
  "Features significant for both abundance and prevalence: ",
  paste(
    intersect(abun_leachate$feature,
              prev_leachate$feature),
    collapse = ", "
  )
)
```

## Methanogenesis

### PWY-1882

[*superpathway of C1 compounds oxidation to
CO2*](https://metacyc.org/pathway?orgid=META&id=PWY-1882)

Pathway Summary Methylotrophic bacteria are aerobic bacteria that
utilize one-carbon compounds more reduced than formate as sources of
carbon and energy, and assimilate formaldehyde as a major source of
cellular carbon \[Hanson96\]. Methylotrophic bacteria utilize a variety
of different one-carbon compounds including methane (see methane
oxidation to methanol I), methanol, methylated amines, halomethanes and
methylated compounds containing sulfur. The common product of these
oxidation reactions is formaldehyde, which is either oxidized to CO2 via
formate, or assimilated (see formaldehyde assimilation and formaldehyde
oxidation pathways).

Note: This is a chimeric pathway, comprising reactions from multiple
organisms, and typically will not occur in its entirety in a single
organism. The taxa listed here are likely to catalyze only subsets of
the reactions depicted in this pathway.

Please note: This superpathway integrates several base pathways that are
found in different methylotrophic bacteria, for the purpose of
displaying the different routes found in the biosphere in a single
general diagram. There may be no single organism that possesses all of
these pathways.

For more information about the pathways and enzymes shown in here,
please see the individual pathways that make up this superpathway.

### PWY-5209

[*methyl-coenzyme M oxidation to CO2
I*](https://metacyc.org/pathway?orgid=META&id=PWY-5209)

Pathway Summary General Background Methanogenesis, the biological
production of methane, is an anaerobic respiration process carried out
by the methanogens, a group of microorganisms belonging to the Archaea
domain. These organisms account for most of the biogenic methane
production, which is estimated at 5x1014 g of methane per year.

Methanogenic pathways utilize a small group of fermentation products
formed by other anaerobes as electron donors and acceptors. Typical
electron donors are H2, formate, and one of several alcohols, such as
ethanol, propan-2-ol, butan-2-ol, or 3-methylbutanol. Typical electron
acceptors are C-1 substrates (carbon-containing compounds that lack
carbon-carbon bonds) such as CO2, methanol, and several different
methylamines and methylsulfides. Acetate is a special methanogenic
substrate, as it serves as both the electron donor and the electron
acceptor after being split into two parts (see methanogenesis from
acetate). Most of the methane in nature originates from acetate.

Four main types of methanogenic pathways have been characterized: the
pathway from H2 and CO2 (see methanogenesis from H2 and CO2), the
aceticlastic pathway from acetate (see methanogenesis from acetate),
methanogenesis from methylated compounds, and methanogenesis from
methoxylated aromatic compounds (see methanogenesis from methoxylated
aromatic compounds). All pathways involve the transfer of a methyl group
from the terminal acceptor to coenzyme M, forming methyl-CoM, which is
then disproportionated into methane and CO2: One in four methyl-CoM
molecules is oxidized to CO2 (see methyl-coenzyme M oxidation to CO2 I),
providing the six electrons that are required for the reduction of three
methyl-CoM molecules to methane (see methyl-coenzyme M reduction to
methane) \[Keltjens93, Pritchett05\].

Methanogenic pathways do not afford substrate level phosphorylation.
Moreover, in the aceticlastic pathway an ATP molecule is spent for
acetyl-CoA formation. Instead, ATP is formed by a chemiosmotic
mechanism. The CoB-CoM heterodisulfide, which is produced during methane
formation (see methyl-coenzyme M reduction to methane), is restored into
coenzyme B and coenzyme M by the action of two membrane-bound
dehydrogenases (F420H2:methanophenazine dehydrogenase and F420
non-reducing hydrogenase I), both of which translocate protons across
the memberane (see coenzyme B/coenzyme M regeneration I
(methanophenazine-dependent)). The translocation results in a proton
gradient that energizes ATP synthase enzymes.

About This Pathway

This pathway covers the anaerobic oxidation of methyl-CoM to CO2. It
should be noted that most of the enzymes in this pathway catalyze
reversible reactions, and operate in different directions in the two
branchs of the pathway.

One of the differences between the methanogens and other methylotrophic
bacteria is that the methanogens often have modified versions of the
tetrahydromethanopterin cofactor and its derivatives. Some organisms
produce only tetrahydromethanopterin (see methyl-coenzyme M oxidation to
CO2 II). Some methanogens that belong to the Methanosarcinales and
Methanococcales orders form tetrahydrosarcinapterin, which is generated
from tetrahydromethanopterin by the addition of a glutamate residue
\[vanBeelen84\] (this pathway). Members of the genus Methanogenium
contain tatiopterin, which is different from tetrahydrosarcinapterin in
having an additional aspartate in the side chain of the molecule, and in
not having the 7-methyl group in the pterin moiety
\[RaemakersFranke90\].

## CO2 Fixation

### PWY-7784

[*reductive acetyl coenzyme A pathway II (autotrophic
methanogens)*](https://metacyc.org/pathway?orgid=META&id=PWY-7784)

Some taxa known to possess this pathway include : [Archaeoglobus
fulgidus](https://metacyc.org/META/NEW-IMAGE?object=TAX-2234),
[Methanosarcina
thermophila](https://metacyc.org/META/NEW-IMAGE?object=TAX-2210),
[Methanothermobacter
thermautotrophicus](https://metacyc.org/META/NEW-IMAGE?object=TAX-145262)

Expected Taxonomic Range:
[Archaea](https://metacyc.org/META/NEW-IMAGE?object=TAX-2157),
[Methanobacteria](https://metacyc.org/META/NEW-IMAGE?object=TAX-183925),
[Methanococci](https://metacyc.org/META/NEW-IMAGE?object=TAX-183939),
[Methanomicrobia](https://metacyc.org/META/NEW-IMAGE?object=TAX-224756),
[Methanopyri](https://metacyc.org/META/NEW-IMAGE?object=TAX-183988)

**Pathway Summary**

This pathway describes autotrophic production of
[acetyl-CoA](https://metacyc.org/compound?orgid=META&id=ACETYL-COA) from
two [CO~2~](https://metacyc.org/compound?orgid=META&id=CARBON-DIOXIDE)
molecules. The pathway was originally documented in homoacetogenic
[*Clostridia*](https://metacyc.org/META/NEW-IMAGE?object=TAX-186801)
\[[Jansen82](https://metacyc.org/META/reference.html?type=CITATION-REFERENCE&object=%5bJansen82%5d)\]
(see [reductive acetyl coenzyme A pathway I (homoacetogenic
bacteria)](https://metacyc.org/pathway?orgid=META&id=CODH-PWY)). In that
pathway the methyl carbon of
[acetyl-CoA](https://metacyc.org/compound?orgid=META&id=ACETYL-COA) is
derived from CO~2~ via the reductions of the tetrahydrofolate route,
while the carbonyl group is obtained via reduction of a second CO~2~ via
CO.

In methanogens, a modified pathway exists, where the methyl group of
[acetyl-CoA](https://metacyc.org/compound?orgid=META&id=ACETYL-COA) is
derived from CO~2~ via the tetrahydromethanopterin route, which is also
a part of the of methanogenic pathway (see [methanogenesis from H~2~ and
CO~2~](https://metacyc.org/pathway?orgid=META&id=METHANOGENESIS-PWY)).

The evidence for the existence of this pathway in methanogens derives
from three observations:

1\. No other known autotrophic pathways exist in these organisms
\[[Zeikus77](http://www.ncbi.nlm.nih.gov/pubmed/914779),
[Daniels78](http://www.ncbi.nlm.nih.gov/pubmed/101522),
[Weimer78](http://www.ncbi.nlm.nih.gov/pubmed/718369),
[Shieh87](http://www.ncbi.nlm.nih.gov/pubmed/3667534)\].

2\. Isotopic labeling of whole cells show that acetate or acetyl-CoA is
the first product of CO~2~ fixation
\[[Fuchs80](http://www.springerlink.com/content/j333123631m75100/fulltext.pdf),
[Ruehlemann85](http://link.springer.com/10.1007/BF00428856)\].

3\. The enzymatic activities associated with this pathway were detected
in cell extracts
\[[Diekert78](http://www.ncbi.nlm.nih.gov/pubmed/711675),
[Drake81](http://www.ncbi.nlm.nih.gov/pubmed/7287757),
[Stupperich83](http://www.ncbi.nlm.nih.gov/pubmed/6840273),
[Lange87](http://www.ncbi.nlm.nih.gov/pubmed/3102234)\].

The pathway starts like the methanogenic pathway with activation of one
molecule of
[CO~2~](https://metacyc.org/compound?orgid=META&id=CARBON-DIOXIDE) by
the unique cofactor
[methanofuran](https://metacyc.org/compound?orgid=META&id=Methanofurans),
resulting in the formation of [a
formylmethanofuran](https://metacyc.org/compound?orgid=META&id=Formyl-methanofurans).
The formyl group is then transferred to another cofactor,
[tetrahydromethanopterin](https://metacyc.org/compound?orgid=META&id=THMPT).
A succession of transformations, catalyzed by
[methenyltetrahydromethanopterin
cyclohydrolase](https://metacyc.org/gene?orgid=META&id=MCHMAUTO-MONOMER),
[H~2~-forming methylene-H~4~MPT
dehydrogenase](https://metacyc.org/gene?orgid=META&id=HMDMAUTO-MONOMER),
and finally [F~420~-dependent methylene-H~4~MPT
reductase](https://metacyc.org/gene?orgid=META&id=MERMAUTO-MONOMER),
which depends on the methanogenic cofactor [a factor
420](https://metacyc.org/compound?orgid=META&id=Factor-420), results in
the formation of
[5-methyltetrahydromethanopterin](https://metacyc.org/compound?orgid=META&id=METHYL-THMPT).

This last compound serves as the branch point between methanogenesis and
acyl-CoA biosynthesis. During methanogenesis, it serves its methyl group
to [coenzyme M](https://metacyc.org/compound?orgid=META&id=CoM) in a
reaction catalyzed by [EC 7.2.1.4, tetrahydromethanopterin
*S*-methyltransferase](https://metacyc.org/META/NEW-IMAGE?type=EC-NUMBER&object=EC-7.2.1.4).
However, during acetyl-CoA biosynthesis it transfers the methyl group to
a corrinoid protein that forms a part of acetyl-CoA synthase. This
transfer is likely catalyzed by a homolog of [EC 2.1.1.258,
5-methyltetrahydrofolate—corrinoid/iron-sulfur protein
*Co*-methyltransferase](https://metacyc.org/META/NEW-IMAGE?type=EC-NUMBER&object=EC-2.1.1.258).
This methyl group then reacts with the CO formed by [EC 1.2.7.4,
anaerobic carbon monoxide
dehydrogenase](https://metacyc.org/META/NEW-IMAGE?type=EC-NUMBER&object=EC-1.2.7.4),
producing
[acetyl-CoA](https://metacyc.org/compound?orgid=META&id=ACETYL-COA).

[Acetyl-CoA](https://metacyc.org/compound?orgid=META&id=ACETYL-COA) is
used as the precursor of carbohydrate synthesis, as described in
[gluconeogenesis II (*Methanobacterium
thermoautotrophicum*)](https://metacyc.org/pathway?orgid=META&id=PWY-6142).

It should be noted that the non-methanogenic archaeon [*Archaeoglobus
fulgidus*](https://metacyc.org/META/NEW-IMAGE?object=TAX-2234) also
utilizes this pathway, but converts the formed acetyl-CoA to acetate in
a reaction that produces ATP. Acetate is the sole product produced by
that organism in the absence of sulfate
\[[Musfeldt02](http://www.ncbi.nlm.nih.gov/pubmed/11790732),
[IngramSmith07](http://www.ncbi.nlm.nih.gov/pubmed/17350930),
[Henstra07](http://www.ncbi.nlm.nih.gov/pubmed/17564616),
[Ferry15](http://www.ncbi.nlm.nih.gov/pubmed/26068860)\].

## Biosynthesis carboxylic acid degradation/biosynthesis

### PWY-7039

[phosphatidate metabolism, as a signaling
molecule](https://metacyc.org/pathway?orgid=META&id=PWY-7039#SUMMARY)

Some taxa known to possess this pathway include : [Arabidopsis thaliana
col](https://metacyc.org/META/NEW-IMAGE?object=ORG-5993)[![Inferred from
experiment](https://metacyc.org/EV-EXP.png){alt="Inferred from experiment"}](javascript:popupUrlContent('/META/reference.html?type=EVIDENCE-GLYPH&object=((PWY-7039+SPECIES+ORG-5993)+%2216081412:EV-EXP:3544209407:zhang%22)%27,%20400,%20400,%20%27Evidence%27,%20this))
\[[GomezMerino05](http://www.ncbi.nlm.nih.gov/pubmed/16081412)\],
[Catharanthus
roseus](https://metacyc.org/META/NEW-IMAGE?object=TAX-4058), [Triticum
aestivum](https://metacyc.org/META/NEW-IMAGE?object=TAX-4565)

Expected Taxonomic Range:
[Viridiplantae](https://metacyc.org/META/NEW-IMAGE?object=TAX-33090)

**Pathway Summary**

[1,2-Diacyl-*sn*-glycerol-3-phosphate](https://metacyc.org/compound?orgid=META&id=L-PHOSPHATIDATE),
often referred to as phosphatidate (PA), is a class of compounds
consisting of a
[glycerol](https://metacyc.org/compound?orgid=META&id=GLYCEROL)
backbone, a (usually) saturated fatty acid bonded to carbon 1, a
(usually) unsaturated fatty acid bonded to carbon 2, and a phosphate
group esterified to carbon 3.

PA is an intermediate in structural lipid biosynthesis (see
[superpathway of phospholipid biosynthesis II
(plants)](https://metacyc.org/pathway?orgid=META&id=PHOSLIPSYN2-PWY) and
[diacylglycerol and triacylglycerol
biosynthesis](https://metacyc.org/pathway?orgid=META&id=TRIGLSYN-PWY)),
and an important second messenger. In plants, PA is rapidly and
transiently generated in response to a number of biotic (pathogens) and
abiotic (such as cold and salt stress) stimulates (reviewed in
\[[Testerink05](http://www.ncbi.nlm.nih.gov/pubmed/16023886)\]). Using a
so-called differential labeling method, it was shown that there are two
pathways, the phospholipase C and diacylglycerol kinase pathway (PLC
pathway), and the phospholipase D pathway (PLD pathway), that generate
PA as a signaling molecule in response to environmental signals
(reviewed in
\[[Arisz09](http://www.ncbi.nlm.nih.gov/pubmed/19394438)\]). In the PLC
pathway, phosphatidylinositol-4,5-bisphosphate, derived from
phosphatidylinositol, is converted to diacylglycerol which is rapidly
phosphorylated to PA. In the PLD pathway, PA is directly formed from
structural lipid phosphatidylcholine. The two pathways are
differentially activated in response to different stimulates
\[[Testerink05](http://www.ncbi.nlm.nih.gov/pubmed/16023886)\]. Once
formed, PA can be further converted to diacylglycerol pyrophosphate
(DGPP) by PA kinase (reviewed in
\[[vanSchooten06](http://www.ncbi.nlm.nih.gov/pubmed/16469533)\]). This
is a possible mechanism in PA attenuation. Interestingly, mammals do not
seem to have PA kinase activity

### PWY-6165

\[*chorismate biosynthesis II (archaea)*\](chorismate biosynthesis II
(archaea)

Some taxa known to possess this pathway include : [Methanocaldococcus
jannaschii](https://metacyc.org/META/NEW-IMAGE?object=TAX-2190),
[Methanococcus
maripaludis](https://metacyc.org/META/NEW-IMAGE?object=TAX-39152)

Expected Taxonomic Range:
[Archaea](https://metacyc.org/META/NEW-IMAGE?object=TAX-2157)

**Pathway Summary**

[Chorismate](https://metacyc.org/compound?orgid=META&id=CHORISMATE) is
an important intermediate that leads to the biosynthesis of several
essential metabolites, including aromatic amino acids, vitamins E and K,
ubiquinone and certain siderophores
\[[Bentley90](http://www.ncbi.nlm.nih.gov/pubmed/2279393)\].
[chorismate](https://metacyc.org/compound?orgid=META&id=CHORISMATE) is
synthesized in 5 steps from
[3-dehydroquinate](https://metacyc.org/compound?orgid=META&id=DEHYDROQUINATE)
(see [chorismate biosynthesis from
3-dehydroquinate](https://metacyc.org/pathway?orgid=META&id=PWY-6163)).

In most organisms,
[3-dehydroquinate](https://metacyc.org/compound?orgid=META&id=DEHYDROQUINATE)
is synthesized from [D-erythrose
4-phosphate](https://metacyc.org/compound?orgid=META&id=ERYTHROSE-4P) in
two steps (see [3-dehydroquinate biosynthesis
I](https://metacyc.org/pathway?orgid=META&id=PWY-6164)). However, the
genomes of the archaea contain no orthologs for the genes that encode
these two enzymes. Instead, at least the euryarchaeota appear to utilize
an alternative pathway in which
[3-dehydroquinate](https://metacyc.org/compound?orgid=META&id=DEHYDROQUINATE)
is synthesized from [6-deoxy-5-ketofructose
1-phosphate](https://metacyc.org/compound?orgid=META&id=CPD-10791) and
[L-aspartate
4-semialdehyde](https://metacyc.org/compound?orgid=META&id=L-ASPARTATE-SEMIALDEHYDE)
\[[White04](http://www.ncbi.nlm.nih.gov/pubmed/15182204)\]. These two
compounds are first condensed to form
[2-amino-3,7-dideoxy-D-threo-hept-6-ulosonate](https://metacyc.org/compound?orgid=META&id=CPD-10792),
which cyclizes to
[3-dehydroquinate](https://metacyc.org/compound?orgid=META&id=DEHYDROQUINATE).
From
[3-dehydroquinate](https://metacyc.org/compound?orgid=META&id=DEHYDROQUINATE)
and on to
[chorismate](https://metacyc.org/compound?orgid=META&id=CHORISMATE), the
archaeal pathway appears to be identical to the bacterial pathway
\[[Porat04](http://www.ncbi.nlm.nih.gov/pubmed/15262931),
[Porat06](http://www.ncbi.nlm.nih.gov/pubmed/17010158)\].

It has been shown that the source for [6-deoxy-5-ketofructose
1-phosphate](https://metacyc.org/compound?orgid=META&id=CPD-10791) is
glycolytic [β-D-fructofuranose
1,6-bisphosphate](https://metacyc.org/compound?orgid=META&id=FRUCTOSE-16-DIPHOSPHATE),
which is condensed with
[methylglyoxal](https://metacyc.org/compound?orgid=META&id=METHYL-GLYOXAL)
\[[White06](http://www.ncbi.nlm.nih.gov/pubmed/17014089)\]. The latter
is believed to be produced by a spontaneous reaction from
[D-glyceraldehyde
3-phosphate](https://metacyc.org/compound?orgid=META&id=GAP), which is
known to be unstable at temperatures above 60°C, and to decompose to
[methylglyoxal](https://metacyc.org/compound?orgid=META&id=METHYL-GLYOXAL)
\[[Richard91](http://www.ncbi.nlm.nih.gov/pubmed/2021650),
[White06](http://www.ncbi.nlm.nih.gov/pubmed/17014089)\].

### PWY-6160

[*3-dehydroquinate biosynthesis II
(archaea)*](https://metacyc.org/pathway?orgid=META&id=PWY-6160)

Some taxa known to possess this pathway include : Methanocaldococcus
jannaschii, Methanococcus maripaludis

Expected Taxonomic Range: Archaea

Pathway Summary chorismate is an important intermediate that leads to
the biosyntrhesis of several essential metabolites, including aromatic
amino acids, 4-aminobenzoate (PABA), vitamins E and K, ubiquinone and
certain siderophores \[Bentley90\]. chorismate is synthesized in 5 steps
from 3-dehydroquinate (see chorismate biosynthesis from
3-dehydroquinate). In most organisms, 3-dehydroquinate is synthesized
from D-erythrose 4-phosphate in two steps (see 3-dehydroquinate
biosynthesis I). However, the genomes of the archaea contain no
orthologs for the genes that encode these first two steps. Instead,
archaeabacteria appear to utilize an alternative pathway in which
3-dehydroquinate is synthesized from 6-deoxy-5-ketofructose 1-phosphate
and L-aspartate 4-semialdehyde \[White04\]. These two compounds are
first condensed to form 2-amino-3,7-dideoxy-D-threo-hept-6-ulosonate,
which cyclizes to 3-dehydroquinate. From 3-dehydroquinate and on to
chorismate, the archaeal pathway appears to be identical to the
bacterial pathway \[Porat04, Porat06\].

It has been shown that the source for 6-deoxy-5-ketofructose 1-phosphate
is glycolytic β-D-fructofuranose 1,6-bisphosphate, which is condensed
with methylglyoxal \[White06\]. The latter is believed to be produced by
a spontaneous reaction from D-glyceraldehyde 3-phosphate, which is known
to be unstable at temperatures above 60°C and to decompose to
methylglyoxal \[Richard91, White06\].

The product of this pathwy, 3-dehydroquinate, not only feeds the
shikimate pathway, but is also the source for 4-aminobenzoate, an
important precursor for the synthesis of tetrahydromethanopterin
\[Porat06\].

### PWY0-301

[*L-ascorbate degradation I (bacterial,
anaerobic)*](https://metacyc.org/pathway?orgid=META&id=PWY0-301) Pathway
Summary General Background L-ascorbate, also known as vitamin C, fulfils
multiple essential roles in both plants and animals. Being a strong
reducing agent, it functions as an antioxidant and a redox buffer. It is
also a cofactor for several enzymes, which are involved in many
important pathways, including collagen hydroxylation, carnitine
biosynthesis, norepinephrine biosynthesis, and hormone and tyrosine
metabolism. In plants L-ascorbate is also implicated in defense against
pathogens and in control of plant growth and development. A significant
proportion of a plant's ascorbate is found in the apoplast (the aqueous
solution permeating the cell walls) \[Green05\].

Under aerobic conditions L-ascorbate is oxidized in cells to
dehydroascorbate (via the radical monodehydroascorbate radical), which
can be recycled back to ascorbate by the ascorbate glutathione cycle.
However, once formed, dehydroascorbate can be further broken down in
vivo by irreversible reactions, escaping the ascorbate glutathione
cycle.

Several pathways for the irreversible catabolism of ascorbate have been
described. Facultatively aerobic bacteria such as Escherichia coli and
Klebsiella pneumoniae degrade L-ascorbate by different pathways under
aerobic and anaerobic conditions (see L-ascorbate degradation II
(bacterial, aerobic) and L-ascorbate degradation I (bacterial,
anaerobic)). The anaerobic pathway begins with phosphorylation of
ascorbate (mediated by a PTS-type transporter), while the aerobic
pathway proceeds via 2,3-didehydro-L-gulonate. Both pathways produce
D-xylulose 5-phosphate, a centeral metabolite that is fed into the
pentose phosphate pathway \[Campos08\].

Plants from the Vitaceae family (e.g. grapes) metabolize ascorbate to
L-tartrate via the intermediates 2-keto-L-gulonate and L-idonate (see
pathway L-ascorbate degradation IV). The tartrate skeleton is derived
from carbons 1-4 of L-ascorbate, indicating a cleavage between carbons 4
and 5 \[Loewus99, DeBolt06\].

The geraniaceous plant Pelargonium crispum metabolizes ascorbate to
L-tartrate and oxalate via a different pathway, with L-threonate, rather
than L-idonate, as an intermediate (see pathway L-ascorbate degradation
III). In this case the tartrate skeleton is derived from carbons 3-6 of
L-ascorbate, indicating a cleavage between carbons 2 and 3 \[Loewus99,
Franceschi05\]. Grapes are also known to accumulate oxalate
\[DeBolt04\], and thus may be using both pathways to generate tartrate.

About This Pathway

Escherichia coli is able to utilize L-ascorbate (vitamin C) as the sole
source of carbon under anaerobic conditions \[Yew02\].

The ula regulon encodes all of the proteins involved in this pathway and
is formed by two operons, ulaG and ulaA-F. ulaG encodes an
L-ascorbate-6-phosphate lactonase, and ulaA-F encodes the three
components of the L-ascorbate specific PTS enzyme II (L-ascorbate
specific PTS enzyme IIC component, L-ascorbate specific PTS enzyme IIB
component, and L-ascorbate specific PTS enzyme IIA component) and three
catabolic enzymes (UlaDEF).

L-ascorbate is imported and converted to L-ascorbate 6-phosphate by the
L-ascorbate specific PTS enzyme II \[Zhang03e\]. The intracellular
L-ascorbate-6-phosphate is subsequently metabolized by UlaG, UlaD, UlaE
and UlaF to D-D-xylulose 5-phosphate, which can enter central metabolism
via the non-oxidative branch of the pentose phosphate pathway \[Yew02\].

Expression of the ula regulon is regulated by the L-ascorbate
6-phosphate-binding repressor UlaR and by cAMP-CRP \[Campos04,
Garces08\].

Under aerobic conditions an additional operon, yiaK-S, is required to
catabolize L-ascorbate (see L-ascorbate degradation II (bacterial,
aerobic)).

## Precursor metabolites

### PWY-5741

[ethylmalonyl-CoA
pathway](https://metacyc.org/pathway?orgid=META&id=PWY-5741#SUMMARY)

Some taxa known to possess this pathway include : Cereibacter
sphaeroides, Methylorubrum extorquens AM1, Paracoccus versutus,
Rhodobacter capsulatus, Rhodospirillum rubrum, Streptomyces coelicolor

Expected Taxonomic Range: Bacteria <bacteria>

Pathway Summary Some organisms utilize organic substrates (such as fatty
acids, alcohols, esters, as well as waxes, alkenes, and methylated
compounds) that are metabolized via acetyl-CoA. These organisms need to
find a way to convert acetyl-CoA, a 2-carbon compound, to a 4-carbon
compound that could feed their anaplerotic reactions (reactions that
form metabolic intermediates for biosynthesis). A common solution to the
problem is the glyoxylate cycle, a modified version of the TCA cycle
that bypasses those steps in the cycle that lead to a loss of CO2.
Acetyl-CoA enters the cycle at two steps, and since no carbon escapes it
in the form of CO2, the output is the 4-carbon compound succinate.

However, many bacteria are known to not contain EC 4.1.3.1, isocitrate
lyase, the key enzyme of the glyoxylate cycle. These organisms include
many purple nonsulfur bacteria, such as Cereibacter sphaeroides and
Rhodospirillum rubrum, and other α-proteobacteria, such as the
methylotroph Methylorubrum extorquens AM1 and the facultative
denitrifier Paracoccus versutus. These organisms require a different
solution to the problem.

One such solution, the ethylmalonyl-CoA pathway, has been discovered in
Cereibacter sphaeroides. In this pathway a C4-compound, acetoacetyl-CoA,
derived from two acetyl-CoA molecules, is converted to the C5-compound
2-methylfumaryl-CoA \[Alber06\]. (2R,3S)-β-methylmalyl-CoA, formed by
hydration of 2-methylfumaryl-CoA, is cleaved to glyoxylate and
propanoyl-CoA. Condensation of glyoxylate and another molecule of
acetyl-CoA yields (S)-malate, while propionyl-CoA is carboxylated to
succinate via a dedicated pathway.

The key enzyme of the pathway is crotonyl-CoA carboxylase/reductase,
which simultaneously carboxylates and reduces the 4-carbon compound
crotonyl-CoA, forming the 5-carbon compound (2S)-ethylmalonyl-CoA.

Another pathway that addresses the same needs, the methylaspartate
cycle, operates in Halobacteria \[Khomyakova11\].)

### PWY-5109

[*propanoate fermentation to
2-*methylbutanoate](https://metacyc.org/pathway?orgid=META&id=PWY-5109)

ome taxa known to possess this pathway include : [Ascaris
lumbricoides](https://metacyc.org/META/NEW-IMAGE?object=TAX-6252),
[Ascaris suum](https://metacyc.org/META/NEW-IMAGE?object=TAX-6253)

Expected Taxonomic Range:
[Metazoa](https://metacyc.org/META/NEW-IMAGE?object=TAX-33208)

**Pathway Summary**

[*Ascaris
lumbricoides*](https://metacyc.org/META/NEW-IMAGE?object=TAX-6252) and
[*Ascaris suum*](https://metacyc.org/META/NEW-IMAGE?object=TAX-6253) are
parasitic intestinal helminths whose metabolism is predominantly
anaerobic \[[Epps50](http://www.ncbi.nlm.nih.gov/pubmed/14774531)\]. The
organisms ferment glucose to
[CO~2~](https://metacyc.org/compound?orgid=META&id=CARBON-DIOXIDE) and a
mixture of products. The major product (about 20%) is
[2-methylbutanoate](https://metacyc.org/compound?orgid=META&id=CPD-7076)
\[[Bueding51](http://www.ncbi.nlm.nih.gov/pubmed/14907729)\].

Saz and Weil have found that the precursors for
[2-methylbutanoate](https://metacyc.org/compound?orgid=META&id=CPD-7076)
in [*Ascaris
lumbricoides*](https://metacyc.org/META/NEW-IMAGE?object=TAX-6252) are
[acetate](https://metacyc.org/compound?orgid=META&id=ACET) and
[propanoate](https://metacyc.org/compound?orgid=META&id=PROPIONATE)
\[[Saz60](http://www.ncbi.nlm.nih.gov/pubmed/14442156)\], which are
produced by the fermentation of either glucose or lactate
\[[Saz59](http://www.ncbi.nlm.nih.gov/pubmed/13673003),
[Saz60](http://www.ncbi.nlm.nih.gov/pubmed/14442156)\]. The enzymatic
steps involved in this process have not been characterized in detail,
but the authors suggested that the pathway consists of the steps
illustrated in here, which are based on the labeling patterns observed
in the study, and on the presence of
[tiglate](https://metacyc.org/compound?orgid=META&id=CPD-7077) in
[*Ascaris*](https://metacyc.org/META/NEW-IMAGE?object=TAX-6251) tissues
\[[Bueding53](http://www.ncbi.nlm.nih.gov/pubmed/13061475)\].

The authors suggest that it is likely that this sequence of reactions
requires the coenzyme A derivatives, similar to the formation of
[butanoate](https://metacyc.org/compound?orgid=META&id=BUTYRIC_ACID)
from [acetate](https://metacyc.org/compound?orgid=META&id=ACET) (see
[pyruvate fermentation to
butanoate](https://metacyc.org/pathway?orgid=META&id=CENTFERM-PWY)).

In subsequent work Saz and Weil also showed that two molecules of
[propanoyl-CoA](https://metacyc.org/compound?orgid=META&id=PROPIONYL-COA)
may be condensed to form
[2-methylpentanoate](https://metacyc.org/compound?orgid=META&id=CPD-15866)
in a similar series of reactions
\[[Saz62](http://www.ncbi.nlm.nih.gov/pubmed/14497735)\]. A putative
acyl-CoA transferase produces the free fatty acids. Both
[2-methylbutanoate](https://metacyc.org/compound?orgid=META&id=CPD-7076)
and
[2-methylpentanoate](https://metacyc.org/compound?orgid=META&id=CPD-15866),
along with [acetate](https://metacyc.org/compound?orgid=META&id=ACET),
[propanoate](https://metacyc.org/compound?orgid=META&id=PROPIONATE) and
[succinate](https://metacyc.org/compound?orgid=META&id=SUC) are excreted
(\[[Komuniecki89](http://www.ncbi.nlm.nih.gov/pubmed/2736251)\] and
reviewed in \[[Muller12](http://www.ncbi.nlm.nih.gov/pubmed/22688819)\])
(see pathway [anaerobic energy metabolism (invertebrates,
mitochondrial)](https://metacyc.org/pathway?orgid=META&id=PWY-7384)).

Unfortunately, most of the enzymes of this pathway have not been
purified yet from
[*Ascaris*](https://metacyc.org/META/NEW-IMAGE?object=TAX-6251).

## Aromatic compound degradation

### PWY-6690

[cinnamate and 3-hydroxycinnamate degradation to
2-hydroxypentadienoate](https://metacyc.org/pathway?orgid=META&id=PWY-6690#SUMMARY)

Some taxa known to possess this pathway include : [Escherichia coli K-12
substr. MG1655](https://metacyc.org/META/NEW-IMAGE?object=TAX-511145),
[Rhodococcus
globerulus](https://metacyc.org/META/NEW-IMAGE?object=TAX-33008)

Expected Taxonomic Range: [Bacteria
\<bacteria\>](https://metacyc.org/META/NEW-IMAGE?object=TAX-2)

**Pathway Summary**

Phenylpropanoid compounds are abundant in natural environments, and can
originate from putrefaction of proteins in soil or as breakdown products
of plant materials such as lignin, various oils, and resins (for the
biosynthesis of phenylpropanoid compounds, please see [Phenylpropanoid
Derivative
Biosynthesis](https://metacyc.org/META/NEW-IMAGE?type=ECOCYC-CLASS&object=PHENYLPROPANOID-SYN)).

Microbial catabolism of phenylpropanoid compounds plays an important
role in the natural degradation of these compounds. In addition, the
degradation of several phenylpropanoid compounds is important for
industrial applications such as wine making, aging, and storage
\[[Cavin97](http://www.ncbi.nlm.nih.gov/pubmed/9143125)\].

The degradation of
[cinnamate](https://metacyc.org/compound?orgid=META&id=CPD-674),
[3-phenylpropanoate](https://metacyc.org/compound?orgid=META&id=3-PHENYLPROPIONATE),
and their hydroxylated derivatives has been reported in several
bacteria, including an Acinetobacter sp.
\[[Dagley65](http://www.ncbi.nlm.nih.gov/pubmed/5881653)\], a
Pseudomonas sp.
\[[Andreoni86](http://www.ncbi.nlm.nih.gov/pubmed/3777934),
[Strickland73](http://www.ncbi.nlm.nih.gov/pubmed/4348920)\], an
Arthrobacter sp.
\[[Strickland73](http://www.ncbi.nlm.nih.gov/pubmed/4348920)\],
[*Escherichia coli*](https://metacyc.org/META/NEW-IMAGE?object=TAX-562)
\[[Burlingame83](http://www.ncbi.nlm.nih.gov/pubmed/6345502)\], and
[*Rhodococcus
globerulus*](https://metacyc.org/META/NEW-IMAGE?object=TAX-33008)
\[[Barnes97](http://www.ncbi.nlm.nih.gov/pubmed/9324265)\].

In [*Escherichia
coli*](https://metacyc.org/META/NEW-IMAGE?object=TAX-562) this pathway
is one of only two aromatic-ring-cleavage pathways that are found in the
organism. The fact that [*Escherichia
coli*](https://metacyc.org/META/NEW-IMAGE?object=TAX-562) can degrade
aromatic acids suggests a wider natural distribution for [*Escherichia
coli*](https://metacyc.org/META/NEW-IMAGE?object=TAX-562) than the
anaerobic environment of the animal gut
\[[Burlingame83](http://www.ncbi.nlm.nih.gov/pubmed/6345502)\]. The
enzymes of this pathway can metabolize
[*trans*-cinnamate](https://metacyc.org/compound?orgid=META&id=CPD-674)
and
[2,3-dihydroxy-*trans*-cinnamate](https://metacyc.org/compound?orgid=META&id=CPD-10796)
as well as
[3-phenylpropanoate](https://metacyc.org/compound?orgid=META&id=3-PHENYLPROPIONATE)
and
[3-(3-hydroxyphenyl)propanoate](https://metacyc.org/compound?orgid=META&id=3-HYDROXYPHENYL-PROPIONATE)
(see [3-phenylpropanoate and 3-(3-hydroxyphenyl)propanoate degradation
to
2-hydroxypentadienoate](https://metacyc.org/pathway?orgid=META&id=HCAMHPDEG-PWY))
\[[Diaz98](http://www.ncbi.nlm.nih.gov/pubmed/9603882)\].

The pathway ultimately yields
[fumarate](https://metacyc.org/compound?orgid=META&id=FUM) and
[2-oxopent-4-enoate](https://metacyc.org/compound?orgid=META&id=OXOPENTENOATE),
which is further degraded to
[pyruvate](https://metacyc.org/compound?orgid=META&id=PYRUVATE) and
[acetyl-CoA](https://metacyc.org/compound?orgid=META&id=ACETYL-COA) as
described in [2-hydroxypenta-2,4-dienoate
degradation](https://metacyc.org/pathway?orgid=META&id=PWY-5162). The
pathway, including the portion that is described in
[2-hydroxypenta-2,4-dienoate
degradation](https://metacyc.org/pathway?orgid=META&id=PWY-5162), has
eight steps, making it one of the longest catabolic sequences known in
[*Escherichia coli*](https://metacyc.org/META/NEW-IMAGE?object=TAX-562)
\[[Burlingame86](http://www.ncbi.nlm.nih.gov/pubmed/3531186),
[Bugg93](http://www.ncbi.nlm.nih.gov/pubmed/8399388),
[Ferrandez97](http://www.ncbi.nlm.nih.gov/pubmed/9098055)\].

### HCAMHPDEG-PWY

[3-phenylpropanoate and 3-(3-hydroxyphenyl)propanoate degradation to
2-hydroxypentadienoate](https://metacyc.org/pathway?orgid=META&id=HCAMHPDEG-PWY#SUMMARY)

Some taxa known to possess this pathway include : [Escherichia coli K-12
substr. MG1655](https://metacyc.org/META/NEW-IMAGE?object=TAX-511145)

Expected Taxonomic Range:
[Pseudomonadota](https://metacyc.org/META/NEW-IMAGE?object=TAX-1224)

**Pathway Summary**

Phenylpropanoid compounds are abundant in natural environments, and can
originate from putrefaction of proteins in soil or as breakdown products
of plants materials such as lignin, various oils, and resins (for the
biosynthesis of phenylpropanoid compounds, please see [Phenylpropanoid
Derivative
Biosynthesis](https://metacyc.org/META/NEW-IMAGE?type=ECOCYC-CLASS&object=PHENYLPROPANOID-SYN)).

Microbial catabolism of phenylpropanoid compounds plays an important
role in the natural degradation of these compounds. In addition, the
degradation of several phenylpropanoid compounds is important for
industrial applications such as wine making, aging, and storage
\[[Cavin97](http://www.ncbi.nlm.nih.gov/pubmed/9143125)\].

This is one of only two aromatic-ring-cleavage pathways found in
[*Escherichia coli*](https://metacyc.org/META/NEW-IMAGE?object=TAX-562).
That [*Escherichia
coli*](https://metacyc.org/META/NEW-IMAGE?object=TAX-562) can degrade
aromatic acids suggests a wider natural distribution for [*Escherichia
coli*](https://metacyc.org/META/NEW-IMAGE?object=TAX-562) than the
anaerobic environment of the animal gut
\[[Burlingame83](http://www.ncbi.nlm.nih.gov/pubmed/6345502)\].

In the utilization of aromatic acids as a carbon and energy source,
[*Escherichia coli*](https://metacyc.org/META/NEW-IMAGE?object=TAX-562)
employs a branched meta-cleavage pathway for the degradation of
[3-phenylpropanoate](https://metacyc.org/compound?orgid=META&id=3-PHENYLPROPIONATE)
and
[3-(3-hydroxyphenyl)propanoate](https://metacyc.org/compound?orgid=META&id=3-HYDROXYPHENYL-PROPIONATE).
Degradation of these two compounds ultimately yields
[succinate](https://metacyc.org/compound?orgid=META&id=SUC),
[acetaldehyde](https://metacyc.org/compound?orgid=META&id=ACETALD) and
[pyruvate](https://metacyc.org/compound?orgid=META&id=PYRUVATE).
Acetaldehyde is further metabolized to yield
[acetyl-CoA](https://metacyc.org/compound?orgid=META&id=ACETYL-COA) (the
second part of this pathway is escribed in [2-hydroxypenta-2,4-dienoate
degradation](https://metacyc.org/pathway?orgid=META&id=PWY-5162)).

The pathway (including the portion that is described in
[2-hydroxypenta-2,4-dienoate
degradation](https://metacyc.org/pathway?orgid=META&id=PWY-5162)) has
eight known steps, which makes it among the longest catabolic sequences
in [*Escherichia
coli*](https://metacyc.org/META/NEW-IMAGE?object=TAX-562)
\[[Burlingame86](http://www.ncbi.nlm.nih.gov/pubmed/3531186),
[Bugg93](http://www.ncbi.nlm.nih.gov/pubmed/8399388),
[Ferrandez97](http://www.ncbi.nlm.nih.gov/pubmed/9098055)\].

The pathway's enzyme are also able to metabolize
[*trans*-cinnamate](https://metacyc.org/compound?orgid=META&id=CPD-674)
and [3-coumarate](https://metacyc.org/compound?orgid=META&id=CPD-10797),
as described in [cinnamate and 3-hydroxycinnamate degradation to
2-hydroxypentadienoate](https://metacyc.org/pathway?orgid=META&id=PWY-6690).

### PWY0-1277

[\*3-phenylpropanoate and 3-(3-hydroxyphenyl)propanoate
degradation\*](https://metacyc.org/pathway?orgid=META&id=PWY0-1277#ONT)

Some taxa known to possess this pathway include : [Escherichia coli K-12
substr. MG1655](https://metacyc.org/META/NEW-IMAGE?object=TAX-511145)

Expected Taxonomic Range:
[Pseudomonadota](https://metacyc.org/META/NEW-IMAGE?object=TAX-1224)

**Pathway Summary**

This is one of only two aromatic-ring-cleavage pathways found in
[*Escherichia coli*](https://metacyc.org/META/NEW-IMAGE?object=TAX-562)
(for the second one, see [phenylacetate degradation I
(aerobic)](https://metacyc.org/pathway?orgid=META&id=PWY0-321)). That
[*Escherichia coli*](https://metacyc.org/META/NEW-IMAGE?object=TAX-562)
can degrade aromatic acids suggests a wider natural distribution for
[*Escherichia coli*](https://metacyc.org/META/NEW-IMAGE?object=TAX-562)
than the anaerobic environment of the animal gut.
\[[Burlingame83](http://www.ncbi.nlm.nih.gov/pubmed/6345502)\] In the
utilization of aromatic acids as a carbon and energy source,
[*Escherichia coli*](https://metacyc.org/META/NEW-IMAGE?object=TAX-562)
employs a branched meta-cleavage pathway for the degradation of
3-phenylpropionate and 3-(3-hydroxyphenyl)propionate. Degradation of
these two compounds ultimately yields succinate, acetaldehyde and
pyruvate. Acetaldehyde is further metabolized to yield acetyl-CoA. The
pathway has eight known steps, which makes it among the longest
catabolic sequences in [*Escherichia
coli*](https://metacyc.org/META/NEW-IMAGE?object=TAX-562).
\[[Burlingame86](http://www.ncbi.nlm.nih.gov/pubmed/3531186),
[Bugg93](http://www.ncbi.nlm.nih.gov/pubmed/8399388),
[Ferrandez97](http://www.ncbi.nlm.nih.gov/pubmed/9098055)\]

### TOLUENE-DEG-3-OH-PWY

[*toluene degradation II (aerobic) (via
4-methylcatechol*](https://metacyc.org/pathway?orgid=META&id=TOLUENE-DEG-3-OH-PWY#SUMMARY)

Expected Taxonomic Range: Pseudomonadota

Pathway Summary Toluene is widely used as an industrial additive and
solvent. Toluene and related aromatic compounds can be degraded by
bacteria, and have been studied in the metabolically versatile genus
Pseudomonas and closely related genera. These studies have been directed
toward bioremediation of environmental pollutants by metabolic
engineering, and the development of syntrophic bacterial consortia
(reviewed in \[Diaz04\]). Aerobic pathways of toluene degradation have
been identified in various species that involve different initial
monooxygenase, or hydroxylating dioxygenase reactions. Several of these
pathways converge in the formation of 3-methylcatechol \[Shields91\].
This compound is a substrate for ring cleavage enzymes, the products of
which are metabolized via a common meta fission pathway, resulting in
the formation of compounds of central metabolism (see this pathway and
pathways toluene degradation to 2-hydroxypentadienoate I (via o-cresol)
and toluene degradation to 2-hydroxypentadienoate (via
toluene-cis-diol)). Toluene degradation in Ralstonia pickettii
(previously known as Pseudomonas pickettii and Burkholderia pickettii)
is controlled by a chromosomal regulon \[Kahng00\]. The initial
hydroxylation reactions are catalyzed by toluene 4-monooxygenase,
followed by meta cleavage of the benzene ring by catechol
2,3-dioxygenase, and subsequent degradation via the meta cleavage
pathway \[Olsen94\].

There is disagreement in the literature over the hydroxylation reactions
of the first enzyme of this pathway. Initial reports suggested that the
first hydroxylation is at the meta position yielding 3-methylphenol as
an intermediate, and the second yields 3-methylcatechol \[Olsen94\].
However, later work suggested that toluene 3-monooxygenase is
predominantly a para cleaving enzyme, and its initial products were
shown in that study to be 90% 4-methylphenol, and only 10%
3-methylphenol. 4-methylphenol was further hydroxylated to
4-methylcatechol \[Fishman04\]. Although further metabolism of
4-methylcatechol was not studied in that report, it has been shown to be
a substrate of catechol 2,3-dioxygenase in Ralstonia pickettii
\[Kukor91\].

Ring fission of 4-methylcatechol, catalyzed by catechol 2,3-dioxygenase,
produces (2Z,4E)-2-hydroxy-5-methyl-6-oxohexa-2,4-dienoate. This
compound is hydrolyzed to (2Z)-2-hydroxyhexa-2,5-dienoate, or its
tautomeric dienol form (this tautomerization occurs spontaneously in
aqueous solution \[Johnson04\]). A second hydrolysis, catalyzed by TbuJ
afford (S)-4-hydroxy-2-oxohexanoate, which is cleaved by the aldolase
TbuK into pyruvate and 1-propanal \[Kukor91\].

The final products of the pathway are pyruvate, which enters central
metabolism, and 1-propanal, which may be converted by EC 1.2.1.87,
propanal dehydrogenase (CoA-propanoylating) to propanoyl-CoA, which can
be converted into the central metabolite succinyl-CoA as described in
propanoyl CoA degradation I.

### PWY-5182

[**toluene degradation II (aerobic) (*via
4-methylcatechol*)**](https://biocyc.org/pathway?orgid=META&id=TOLUENE-DEG-3-OH-PWY&sid=biocyc15-3987718044)

Some taxa known to possess this pathway include : [Ralstonia
pickettii](https://biocyc.org/META/NEW-IMAGE?object=TAX-329)

Expected Taxonomic Range:
[Pseudomonadota](https://biocyc.org/META/NEW-IMAGE?object=TAX-1224)

### **Pathway Summary**

Toluene is widely used as an industrial additive and solvent. Toluene
and related aromatic compounds can be degraded by bacteria, and have
been studied in the metabolically versatile genus
[*Pseudomonas*](https://biocyc.org/META/NEW-IMAGE?object=TAX-286) and
closely related genera. These studies have been directed toward
bioremediation of environmental pollutants by metabolic engineering, and
the development of syntrophic bacterial consortia (reviewed in
\[[Diaz04](http://www.ncbi.nlm.nih.gov/pubmed/15492931)\]). Aerobic
pathways of toluene degradation have been identified in various species
that involve different initial monooxygenase, or hydroxylating
dioxygenase reactions. Several of these pathways converge in the
formation of
[3-methylcatechol](https://biocyc.org/compound?orgid=META&id=CPD-111)
\[[Shields91](http://www.ncbi.nlm.nih.gov/pubmed/1892384)\]. This
compound is a substrate for ring cleavage enzymes, the products of which
are metabolized via a common *meta* fission pathway, resulting in the
formation of compounds of central metabolism (see this pathway and
pathways [toluene degradation to 2-hydroxypentadienoate I (*via*
*o*-cresol)](https://biocyc.org/pathway?orgid=META&id=TOLUENE-DEG-2-OH-PWY)
and [toluene degradation to 2-hydroxypentadienoate (*via*
toluene-*cis*-diol)](https://biocyc.org/pathway?orgid=META&id=TOLUENE-DEG-DIOL-PWY)).

Toluene degradation in [*Ralstonia pickettii* (previously known as
*Pseudomonas pickettii* and *Burkholderia
pickettii*)](https://biocyc.org/META/NEW-IMAGE?object=TAX-329) is
controlled by a chromosomal regulon
\[[Kahng00](http://www.ncbi.nlm.nih.gov/pubmed/10671442)\]. The initial
hydroxylation reactions are catalyzed by [toluene
4-monooxygenase](https://biocyc.org/complex?orgid=META&id=CPLX-6542),
followed by *meta* cleavage of the benzene ring by catechol
2,3-dioxygenase, and subsequent degradation via the *meta* cleavage
pathway \[[Olsen94](http://www.ncbi.nlm.nih.gov/pubmed/8206853)\].

There is disagreement in the literature over the hydroxylation reactions
of the first enzyme of this pathway. Initial reports suggested that the
first hydroxylation is at the *meta* position yielding
[3-methylphenol](https://biocyc.org/compound?orgid=META&id=CPD-112) as
an intermediate, and the second yields
[3-methylcatechol](https://biocyc.org/compound?orgid=META&id=CPD-111)
\[[Olsen94](http://www.ncbi.nlm.nih.gov/pubmed/8206853)\]. However,
later work suggested that toluene 3-monooxygenase is predominantly a
*para* cleaving enzyme, and its initial products were shown in that
study to be 90%
[4-methylphenol](https://biocyc.org/compound?orgid=META&id=CPD-108), and
only 10%
[3-methylphenol](https://biocyc.org/compound?orgid=META&id=CPD-112).
[4-methylphenol](https://biocyc.org/compound?orgid=META&id=CPD-108) was
further hydroxylated to
[4-methylcatechol](https://biocyc.org/compound?orgid=META&id=4-METHYLCATECHOL)
\[[Fishman04](http://www.ncbi.nlm.nih.gov/pubmed/15126473)\]. Although
further metabolism of 4-methylcatechol was not studied in that report,
it has been shown to be a substrate of catechol 2,3-dioxygenase in
[*Ralstonia
pickettii*](https://biocyc.org/META/NEW-IMAGE?object=TAX-329)
\[[Kukor91](http://www.ncbi.nlm.nih.gov/pubmed/1856161)\].

Ring fission of
[4-methylcatechol](https://biocyc.org/compound?orgid=META&id=4-METHYLCATECHOL),
catalyzed by catechol 2,3-dioxygenase, produces
[(2Z,4E)-2-hydroxy-5-methyl-6-oxohexa-2,4-dienoate](https://biocyc.org/compound?orgid=META&id=CPD-8781).
This compound is hydrolyzed to
[(2Z)-2-hydroxyhexa-2,5-dienoate](https://biocyc.org/compound?orgid=META&id=CPD-22385),
or its tautomeric dienol form (this tautomerization occurs spontaneously
in aqueous solution
\[[Johnson04](http://www.ncbi.nlm.nih.gov/pubmed/15301547)\]). A second
hydrolysis, catalyzed by TbuJ afford
[(*S*)-4-hydroxy-2-oxohexanoate](https://biocyc.org/compound?orgid=META&id=CPD-13722),
which is cleaved by the aldolase TbuK into
[pyruvate](https://biocyc.org/compound?orgid=META&id=PYRUVATE) and
[1-propanal](https://biocyc.org/compound?orgid=META&id=CPD-665)
\[[Kukor91](http://www.ncbi.nlm.nih.gov/pubmed/1856161)\].

The final products of the pathway are
[pyruvate](https://biocyc.org/compound?orgid=META&id=PYRUVATE), which
enters central metabolism, and
[1-propanal](https://biocyc.org/compound?orgid=META&id=CPD-665), which
may be converted by [EC 1.2.1.87, propanal dehydrogenase
(CoA-propanoylating)](https://biocyc.org/META/NEW-IMAGE?type=EC-NUMBER&object=EC-1.2.1.87)
to
[propanoyl-CoA](https://biocyc.org/compound?orgid=META&id=PROPIONYL-COA),
which can be converted into the central metabolite
[succinyl-CoA](https://biocyc.org/compound?orgid=META&id=SUC-COA) as
described in [propanoyl CoA degradation
I](https://biocyc.org/pathway?orgid=META&id=PROPIONMET-PWY).

## Chlorinated Compound Degradation

### PCPDEG-PWY

[*pentachlorophenol
degradation*](https://metacyc.org/pathway?orgid=META&id=PCPDEG-PWY)

Expected Taxonomic Range: Bacteria <bacteria>

Pathway Summary Pentachlorophenol (PCP) is a a polychlorinated aromatic
compound that has been released into the environment as a wood
preservative and broad spectrum biocide \[Kaufman77a, Crosby81\]. This
compound is a major environmental pollutant due to its toxicity and
recalcitrance, and it is regulated as one of the priority pollutants by
the U.S. Environmental Protection Agency \[Middaugh94\]. Microorganisms
have been used to remove PCP from the environment \[Miethling96\], and
several aerobic PCP-degrading bacteria have been isolated from
contaminated soils \[Crawford99\]. Sphingobium chlorophenolicum
(previously known as Sphingomonas chlorophenolica) strain ATCC 39723 is
one of the bacteria capable of completely mineralizing PCP \[Saber85\]
The degradation pathway in Sphingobium chlorophenolicum starts with EC
1.14.13.50, pentachlorophenol monooxygenase, an enzyme that catalyzes
the conversion of pentachlorophenol to
2,3,5,6-tetrachloro-1,4-benzoquinone, followed by EC 1.1.1.404,
tetrachlorobenzoquinone reductase, which reduces the latter to
2,3,5,6-tetrachlorohydroquinone. The next enzyme, a glutathione
transferase encoded by the pcpC gene, catalyzes the next two steps,
resulting in 2,6-dichlorohydroquinone. PcpC is susceptible to oxidative
damage, and the damaged PcpC produces glutathionyl (GS) conjugates which
it can't process further. These conjugates can be rescued by the action
of another glutathione transferase encoded by pcpF, which completes the
reduction reaction and removes the glutathione moiety.

The next enzyme, 2,6-dichlorohydroquinone dioxygenase, cleaves the ring,
resulting in the non-aromatic product 2-chloromaleylacetate \[Ohtsubo99,
Lange96, Orser94\]. This compound is dehalogenated in two steps by
maleylacetate reductase, resulting in 3-oxoadipate, a common
intermediate in the degradation of aromatic compounds, which is further
degraded into central metabolism intermediates.

The genes responsible for degradation of pentachlorophenol are found on
a secondary chromosome that encodes primarily genes that appear to be
involved in environmental adaptation \[Copley12\]. The first and third
enzymes in this pathway (pentachlorophenol hydroxylase and
2,6-dichlorohydroquinone dioxygenase) may have originated from enzymes
in a pathway for degradation of a naturally occurring chlorinated
phenol. The second enzyme, tetrachlorohydroquinone reductive
dehalogenase, may have evolved from a maleylacetoacetate isomerase (for
example, see maleylacetoacetate isomerase) which is normally involved in
degradation of tyrosine \[Copley00\]. These three genes were acquired
fairly recently via two different horizontal gene transfer events, while
the genes encoding the final two enzymes were acquired earlier
\[Copley12\].

The pathway is not very effective - pentachlorophenol hydroxylase is
very slow, and tetrachlorohydroquinone reductive dehalogenase is subject
to severe product inhibition, further suggesting that the pathway has
evolved recently \[Copley00\].

### 14DICHLORBENZDEG-PWY

::: callout-note
Aromatic & Chlorinated Compound Degradation !
:::

[*14-dichlorobenzene
degradation*](https://metacyc.org/pathway?orgid=META&id=14DICHLORBENZDEG-PWY)

Expected Taxonomic Range: Pseudomonadota

Pathway Summary 1,4-Dichlorobenzene is a volatile organic compound used
in mothballs and some air fresheners, and a common contaminant of indoor
air. It is also used against termites in soil. It has been classified as
a possible carcinogen and may be related to decreased pulmonary function
in adults \[Elliott08\]. 1,4-Dichlorobenzene has been identified as a
priority pollutant by the US Environmental Protection Agency (EPA).
Chlorinated benzenes are chemically stable in nature, so biological
degradation is the only process by which these compounds are eliminated
\[Spiess95\]. Several bacterial species were reported to catabolize
1,4-dichlorobenzene, including Alcaligenes sp. strain A175 \[Schraa86\],
an unidentified Pseudomonas species \[Spain87\], Pseudomonas sp. P51
\[vanderMeer91\], Xanthobacter flavus 14p1 \[Spiess95\] and other not
well characterized bacteria \[Adebusoye07\].

The best studied case is that of Xanthobacter flavus 14p1, a bacterium
that is able to utilize 1,4-dichlorobenzene as a sole source of carbon
and energy. The degradation pathway in this organism begins with
dioxygenation of 1,4-dichlorobenzene by a chlorobenzene dioxygenase,
followed by chlorobenzene cis-dihydrodiol dehydrogenase, resulting in
3,6-dichlorocatechol. Ring opening then proceeds via a modified ortho
cleavage pathway, yielding 3-oxoadipate which is processed to TCA cycle
intermediates \[Spiess95\]. The intermediates of the pathway up to
2-chloromaleylacetate were isolated and identified by gas
chromatography, confirming the propsed pathway \[Spiess95\]. In
addition, individual enzyme activities have been measured with whole
cells and cell extracts \[Sommer97\].

Even though Xanthobacter flavus 14p1 can not grow with chlorobenzene or
1,3-dichlorobenzene as the sole carbon source, all the enzymes that
participate in this pathway can also accept the respective intermediates
of the degradation of these two compound \[Sommer97\]. In the case of
1,3-dichlorobenzene, lack of growth results probably from failure to
induce the enzymes. Growth with chlorobenzene, on the other hand, does
induce the key enzymes. The authors speculate that lack of growth with
this compound may be the result of toxicity of the one of the
intermediates involved \[Sommer97\]. Lipid metabolism pathways →
membrane stress responses Aromatic degradation pathways → breakdown of
plastic-derived organics Redox / respiration pathways → oxidative stress
handling

### PWY-6084

::: callout-note
Aromatic & Chlorinated Compound Degradation !
:::

[*3,5-dichlorocatechol
degradation*](https://metacyc.org/pathway?orgid=META&id=PWY-6084)

Some taxa known to possess this pathway include : Bradyrhizobium sp.
RD5-C2, Burkholderia sp. RASC, Caballeronia sp. NK8, Cupriavidus
pinatubonensis JMP134, Cupriavidus sp. PS12, Delftia acidovorans,
Pseudomonas knackmussii, Ralstonia eutropha NH9

Expected Taxonomic Range: Bacteria <bacteria>

Pathway Summary Chlorobenzenes are used in the production of pesticides,
dyes, pharmaceuticals, disinfectants, rubbers, plastics, and electric
goods \[Rapp01\]. Their occurrence in the environment is widespread and
they are found in the atmosphere \[Popp00\], water \[Aelion87,
Monferran05\], soil \[Guerin08, Sabljic89\], sediments \[Lee05c\],
vegetables \[Zhang05d\], and animals \[Vorkamp04\]. Chlorobenzenes are
of great concern because of their toxicity, persistence and accumulation
in the food chain, and many of them have been declared priority
pollutants by the US Environmental Protection Agency (EPA). Two
important chlrobenzene compounds, 2,4-dichlorophenoxyacetate and
1,3-dichlorobenzene, are degraded via 3,5-dichlorocatechol. This
catechol is subsequently degraded to 3-oxoadipate in a modified ortho
cleavage pathway, as described in this pathway. Several bacterial
species were reported to possess the pathway, including Cupriavidus
pinatubonensis JMP134 \[Laemmli04\], Cupriavidus sp. PS12
\[Pollmann01\], Pseudomonas knackmussii \[Schwein88\], Bradyrhizobium
sp. RD5-C2 \[Itoh02\], Burkholderia sp. RASC \[Suwa96\], Delftia
acidovorans \[Muller01b\], Caballeronia sp. NK8 \[Liu01a\] and Ralstonia
eutropha NH9 \[Liu05b\]. The enzymes catalyzing this pathway have been
studied and characterized from several organisms \[Schwein88,
vanderMeer91a\]. They are usually encoded by transmissible plasmids,
such as the pJP4 plasmid of Cupriavidus pinatubonensis JMP134 or the
pB13 plasmid of Pseudomonas knackmussii.

The catechol is attacked by a chlorocatechol 1,2-dioxygenase that opens
the ring, generating (2E,4E)-2,4-dichloromuconate. chloromuconate
cycloisomerase I catalyzes a reaction that combines isomerization to a
lactone with the loss of one of the chlorine atoms. In Pseudomonas
knackmussii the lacton that is produced is in the trans form, and is
isomerized to a cis form by a 2-chloro-trans-dienelactone isomerase. It
is not clear if this occurs in other organisms as well \[Schwein88\].
The lactone is then hydrolyzed to 2-chloromaleylacetate by a hydrolase.
Finally, ther last enzyme of the pathway, chloromaleylacetate reductase,
catalyzes two successive reductions, during which the second chlorine
atom is lost. The final product, 3-oxoadipate, is degraded to TCA cycle
intermediates by chromosomally encoded enzymes.

\### PWY-7039

::: callout-note
Biosynthesis fatty acid / lipids phospholipid biosynthesis signaling
:::

[*phosphatidate metabolism, as a signaling
molecule*](https://metacyc.org/pathway?orgid=META&id=PWY-7039)

Expected Taxonomic Range: Viridiplantae

Pathway Summary 1,2-Diacyl-sn-glycerol-3-phosphate, often referred to as
phosphatidate (PA), is a class of compounds consisting of a glycerol
backbone, a (usually) saturated fatty acid bonded to carbon 1, a
(usually) unsaturated fatty acid bonded to carbon 2, and a phosphate
group esterified to carbon 3. PA is an intermediate in structural lipid
biosynthesis (see superpathway of phospholipid biosynthesis II (plants)
and diacylglycerol and triacylglycerol biosynthesis), and an important
second messenger. In plants, PA is rapidly and transiently generated in
response to a number of biotic (pathogens) and abiotic (such as cold and
salt stress) stimulates (reviewed in \[Testerink05\]). Using a so-called
differential labeling method, it was shown that there are two pathways,
the phospholipase C and diacylglycerol kinase pathway (PLC pathway), and
the phospholipase D pathway (PLD pathway), that generate PA as a
signaling molecule in response to environmental signals (reviewed in
\[Arisz09\]). In the PLC pathway, phosphatidylinositol-4,5-bisphosphate,
derived from phosphatidylinositol, is converted to diacylglycerol which
is rapidly phosphorylated to PA. In the PLD pathway, PA is directly
formed from structural lipid phosphatidylcholine. The two pathways are
differentially activated in response to different stimulates
\[Testerink05\]. Once formed, PA can be further converted to
diacylglycerol pyrophosphate (DGPP) by PA kinase (reviewed in
\[vanSchooten06\]). This is a possible mechanism in PA attenuation.
Interestingly, mammals do not seem to have PA kinase activity.

# 000PCA of pathway matrix

> Goal: Perform PCA analysis on pathway abundance data and create an
> informative visualization that includes a scatter plot of the first
> two principal components (PC1 vs PC2) with density plots for both PCs.
> The plot helps to visualize the clustering patterns and distribution
> of samples across different groups.

```{r}
# Perform PCA on pathway abundance data
pca_comp <- prcomp(path, scale. = TRUE)

# Create a data frame for plotting
pca_df <- as.data.frame(pca_comp$x)

# Add metadata to the PCA data frame
pca_df <- cbind(pca_df, meta)

# Set factors 
pca_df <- pca_df %>% 
  mutate(stage = factor(stage, levels = c("cleavage", "prawnchip", "earlygastrula")),
         leachate = factor(leachate, levels = c("control", "low", "mid", "high")))

# Calculate %Var in PC1 and PC2
percentVar <- (pca_comp$sdev^2) / sum(pca_comp$sdev^2) * 100

pc1_var <- round(percentVar[1], 1)
pc2_var <- round(percentVar[2], 1)
```

Labels for grouping variables

```{r}
labs_leachate <- c(control = "0 mg/L (control)",
                   low = "0.01 mg/L (low)",
                   mid = "0.1 mg/L (mid)",
                   high = "1 mg/L (high)")

labs_stage <- c(cleavage = "Cleavage", prawnchip = "Prawn chip", earlygastrula = "Early gastrula")
```

```{r}
# Create a scatter plot of PC1 vs PC2 with density plots
ggplot(pca_df, aes(x = PC1, y = PC2, color = leachate)) +
  geom_point(size = 3) +
  #geom_density_2d() +
  labs(title = "PCA of Pathway Abundance Data",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal() +
  theme(legend.title = element_blank())
```

# ggpicrust2 package for visualizing results

Git repo [here](https://github.com/cafferychen777/ggpicrust2#workflow)
...

```{r ggpicrust2}
library(ggpicrust2)
```

```{r}
results_data_input <- ggpicrust2(data = abundance_data,
                                 metadata = metadata,
                                 group = "your_group_column", # For example dataset, group = "Environment"
                                 pathway = "KO",
                                 daa_method = "LinDA",
                                 ko_to_kegg = TRUE,
                                 order = "pathway_class",
                                 p_values_bar = TRUE,
                                 x_lab = "pathway_name")
```

```{r}
# If you want to analysis the EC. MetaCyc. KO without conversions.

path_rows <- t(path)

metacyc_daa_results_df <- pathway_daa(
                                      abundance = path_rows,
                                      metadata = meta,
                                      group = "group",
                                      daa_method = "DESeq2"
                                    )

metacyc_daa_annotated_results_df <- 
  pathway_annotation(pathway="MetaCyc", 
                     daa_results_df = metacyc_daa_results_df,
                     ko_to_kegg = FALSE)

p <- pathway_errorbar(
  abundance = path_rows,
  daa_results_df = metacyc_daa_annotated_results_df,
  Group = meta$group,
  ko_to_kegg = FALSE,
  p_values_threshold = 0.05,
  order = "group",
  select = NULL,
  p_value_bar = TRUE,
  colors = NULL,
  x_lab = "description"
)

p
```

```{r}
# Filter features with p < 0.05
feature_with_p_0.05 <- metacyc_daa_annotated_results_df %>%
  filter(p_adjust < 0.05)

# Create the heatmap
pathway_heatmap(
  abundance = path_rows %>%
    right_join(
      annotated_metacyc_daa_results_df %>% select(all_of(c("feature","description"))),
      by = c("pathway" = "feature")
    ) %>%
    filter(pathway %in% feature_with_p_0.05$feature) %>%
    select(-"pathway") %>%
    column_to_rownames("description"),
  metadata = meta,
  group = "group"
)
```
